Quantifying microbial fitness in high-throughput experiments

Version 1

Abstract

Few concepts are as central to evolution as is fitness, and yet the quantification of fitness is often ambiguous. In particular, high-throughput experiments to measure mutant fitness in microbes are increasingly common but vary widely in their definitions of fitness, which makes their results difficult to compare. What are the consequences of these different fitness statistics, and is there a best way to quantify fitness in a given context? Here we systematize the set of possible fitness statistics according to the following three choices: 1) the encoding of relative abundance (e.g., transforming by a log or logit function), 2) the time scale over which to measure the change in relative abundance, and 3) the choice of a reference subpopulation for calculating fitness in bulk competition experiments, such as those using DNA-barcoded mutants. We show that these different choices can lead to significantly different interpretations of mutant fitness, affecting the magnitude of fitness effects, the presence of epistasis, and even the fitness ranking across mutants. This can confound predictions for evolutionary dynamics and gene functions. Altogether our results demonstrate the importance of consistent fitness definitions for reproducible results across experiments.

Introduction

Fitness describes the fate of mutations as they arise in a population [1] and serves to define other core evolutionary concepts such as trade-offs [2]. At the same time, fitness is an often confusing term in evolutionary biology [3–5]. It is possible to classify fitness measures by their role in theory [6] and their scale of measurement [7–9], but these arguments offer little guidance on how to measure fitness with microorganisms in practice. In particular for the field of microbial ecology and evolution, empirical fitness measurements allow us to test general evolutionary theory with fast-growing model organisms [10– 15] and can also be used to detect microbial interactions [16, 17], annotate gene function [18, 19], and understand the spread of antibiotic resistance genes [20, 21].

The classic approach to measuring the fitness effects of mutations uses pairwise competition experiments, where the mutant competes with a wild-type in co-culture [22– 24]. This approach is ideal because it closely mimics the dynamics of spontaneous mutations, and thus is typically used in experimental evolution [10, 12, 25–27]. For example, the Long-Term Evolution Experiment (LTEE) has evolved Escherichia coli over tens of thousand of generations and measured the fitness of each evolved lineage in competition with the ancestor [11, 26, 28]. Since measuring pairwise competitions becomes infeasible for a large number of strains, a second approach is to measure properties of genotypes in monoculture, like the growth rate, and combine these measurements for a relative fitness estimate. Monoculture growth curves are straightforward to measure, even for entire libraries of single-gene knock-out mutants [29–31], but empirical tests have shown that the monoculture growth rate is insufficient to predict the mutant fitness [32–34]. Summary statistics, like the area under the curve (AUC), perform better [35–37]. However, this approach breaks down for strains that engage in cross-feeding or toxin production and any attempts to improve monoculture-based fitness estimates require additional experiments [34, 37] and do not go beyond qualitative fitness rankings [37].

A third and more recent approach is to measure relative fitness in bulk competition experiments, based on high-throughput barcode sequencing [13, 38]. For example, transposon-insertion mutagenesis generates libraries of barcoded gene knockouts that can be tracked as they grow in a single batch culture [39]. Estimating the fitness of each barcode allows us to identify genes that are particularly important in the growth environment [18, 19, 40], in the given genetic background [14, 15, 41, 42], and in the context of an ecological community [16, 17, 42, 43]. Bulk fitness estimation also plays a major role to identify beneficial lineages in experimental evolution [13, 44] (not unlike the tracking of viral pathogens in public health [45–47]) and to test their fitness in new environments [48–52]. Many of these studies make different choices about the design of experiments and the calculation of fitness, but without explaining why a certain fitness measure was used or how to compare between data sets.

The inconsistent choices for quantifying microbial fitness make it is difficult to compare fitness across experiments, especially because different groups tend to use different definitions. For example, the LTEE reports relative fitness per generation [11, 26], whereas other evolution experiments exclusively report relative fitness percycle [12, 53]. New techniques, like barcode sequencing, often spark their own fitness metrics [18, 40, 54–56]. It is also not clear what the consequences of this choice are; a different fitness metric might lead to a different ranking of mutants in a genetic screen (and the importance we assign to those genes), besides affecting the quantitative fitness value. The probability of fixation for spontaneous mutations [57, 58], the speed of adaptation [59], and the effect size for gene essentiality experiments [18, 19] all depend on our ability to estimate quantitative fitness values. This raises three key questions: How do these fitness statistics differ, or are some equivalent? Can we say which choices are optimal, in the light of first principles or some practical considerations? And does this tell us how we should design these experiments? Previous work has addressed aspects of these questions [6–9, 60] but the arguments are scattered across disciplines and based on generic models of population growth [6, 7, 9, 61], rather than incorporating the specific dynamics of microbes in laboratory experiments. Here we address these questions from a unified framework, using realistic microbial population dynamics with empirical traits to systematically test different fitness metrics.

Results

Predictive power varies across different relative fitness statistics

While there are other notions of fitness relevant in different evolutionary contexts (Sec. S1, Fig. S1), in this article we focus on relative fitness as this is the most common object of high-throughput laboratory experiments in microbes [18, 26, 40, 55, 56]. To generalize the wide range of relative fitness statistics used in these studies, we define a genotype’s relative fitness as any number that is sufficient to predict the genotype’s relative abundance x(t) over a short-time horizon (Fig. S1). The simplest approach is to take a linear expansion of x(t) and use its slope at time t to predict the change over a time-window Δt:

Under this linear approximation, the slope slinear = dx/dt|t constitutes a relative fitness statistic, since it is sufficient to predict the change in relative abundance. Figure 1A (top panel) shows a schematic trajectory of relative abundance x(t) with two examples of this predictive approach (dashed arrows). As we can see, the naive linear statistic of relative fitness can significantly under- or overestimate the actual change in relative abundance, because x(t) changes nonlinearly in time.

Overview of the choice of encoding and the choice of timescale for quantifying relative fitness.

(A) Example trajectory of relative abundance x (upper panel) for a mutant invading and eventually replacing a wild-type population. The same trajectory is plotted under the encoding log x (middle panel) and the logit-encoding log(x/(1 − x)) (lower panel). (B) The general flow-chart to predict the future relative abundance of a mutant given a relative fitness value sm = dm/dt for some encoding m. The current relative abundance xt is transformed into the new variable mt = m(xt), then projected into the future through a linear extrapolation using sm (upper horizontal arrow) and finally converted back into a frequency xt+..6t using the decoding function m−1. (C) Four scenarios for positive mutant fitness with different underlying population dynamics. For each scenario, we show an example trajectory of absolute abundance (stacked) for the wild-type (dark grey) and mutant population (light grey). Each scenario is mapped as a single-dot onto the fold-change diagram (center plot) and colored areas indicated positive (green area) and negative relative fitness per-cycle (blue area; compare Eq. (8)). (D) Basic constellation for misranking between relative fitness per-cycle and relative fitness per-generation . For a given competition (red dot), misranking occurs in a bow-tie area of the fold-change space (red shade). Any competition in the right half of this area (grey dot) will have higher mutant fitness but lower mutant fitness (right inset). As small plots on the left, we show possible population dynamics that generate this fold-change variation.

As an alternative, we can transform the relative abundance x into a new variable m(x) that improves the quality of prediction (Fig. 1B; compare [62]). We define an encoding as a smooth, strictly-increasing function of relative abundance. This one-to-one mapping allows us to predict the change in relative abundance using a linear expansion of the encoded relative abundance, rather than the relative abundance itself: where m−1 is the inverse of the encoding function and is the relative fitness under the encoding m. Note that the relative fitness of neutral mutations is always zero, independent of the encoding.

For example, Fig. 1A (middle panel) shows the relative abundance under the log-encoding m(x) = log x; this encoding is implicit in typical plots of relative abundance on a logarithmic scale [44]. This leads to an approximately linear trajectory in the beginning, where the prediction quality of the slope of the encoding slog is thus higher than for the slope of relative abundance slinear (compare top and middle panels of Fig. 1A). However, under the log-encoding the prediction quality remains poor at later stages in the trajectory. In the case of Fig. 1A, we see that the ideal encoding is given by m(x) = logit x (bottom panel), where the logit function is defined as

This is ideal because it transforms the nonlinear trajectory of relative abundance x(t) into a linear trajectory that can be fully predicted from the linear expansion in Eq. (2).

Why is the logit function m(x) = logit x the ideal encoding of relative abundance for a fitness statistic in this case? Since we generated the relative abundance x(t) in Fig. 1A using logistic dynamics, the logit function is linearly related to the inverse function of x(t) (Sec. S2). More generally, the ideal encoding for a given relative abundance dynamics x(t) is any linear function of the inverse m(x(t)) = at(x) + b of those dynamics. Note that the ambiguous scale a also includes log and logit encodings with a different base of the logarithm (discussed in Ref. [7]). The inversion of x(t) is equivalent to removing frequency (relative abundance) dependence from the relative fitness statistic slinear [63]. For example, Fig. S2 shows that if we apply the logit-encoding to a trajectory with non-logistic underlying dynamics (e.g., Gompertz model), the logit transform reduces but does not entirely remove the nonlinearity in the relative abundance trajectory (see the mathematical notes by Mallet [64] for more examples). We note that is possible to generalize the concept of encodings to absolute abundance as well (Sec. S3).

Of course, the exact dynamics of microbial populations are generally not known, and if they were known, there would be no need to make predictions using the linear expansion in Eq. (2). In the absence of an exact model, there are mathematical and practical reasons for why the logit function is a sensible encoding for relative abundance. First, logistic dynamics can be interpreted as the lowest-order approximation to more complex dynamics of relative abundance (Sec. S2). Second, the logit encoding is particularly suited to the binomial sampling noise in experimental measurements of relative abundance (Fig. S3). Logit is the designated link function for binomial random variables, meaning that it normalizes the measurement variance across relative abundances (heteroscedasticity), regardless of whether the underlying dynamics are logistic [65–67]. For these reasons, we exclusively focus on the logit-based relative fitness for the remainder of this work.

Relative fitness statistics require a choice of timescale

In practice, the relative abundance of a genotype is only available as a trajectory of discrete time points (see crossmarks in Fig. S1C). To estimate the relative fitness at a given time point, we calculate the finite difference of the encoded abundance between the current time t and the previous time point: where Δt is the time difference between observations. For discretized time steps it is also possible to define a multiplicative fitness that describes the ratio, rather than the difference, of encoded relative abundance between time points (Sec. S4).

In the case of a single mutant competing against a wild-type strain, the estimated relative fitness of the mutant (using Eq. (5)) under the logit encoding is where xmut is the relative abundance of the mutant and xwt = 1 − xmut is the relative abundance of the wild-type. This form of relative fitness is widely used in empirical measurements of microbial fitness [12, 66]. Note that a special property of the logit encoding is that we can rewrite Eq. (6) in terms of the log fold-changes of the mutant and wild-type strains: where the log fold-change of each strain is LFC = log N (t)/N (t Δt) and N is the biomass of the strain. The LFC is sometimes also referred to as a Malthusian parameter [9, 26, 68].

A key element of this estimate of relative fitness (Eq. (5)) is the time interval Δt. There are three common ways to choose a time interval Δt in empirical measurements of microbial fitness. The simplest is to use a fixed clock time between measurements (e.g., one day). However, many microbial populations, especially in laboratory experiments [26, 69] but also in some natural environments [70–72], grow in discrete growth cycles dictated by pulses of nutrients (batch culture). These cycles define an intrinsic timescale of population dynamics and also determine the time point of sampling. In this case, it is convenient to quantify relative fitness per cycle where we have chosen the logit-encoding and Δt = 1 growth cycle in Eq. (5).

Another important timescale for microbial populations is the generation time. In general, the relative fitness per cycle (Eq. (8)) may depend on the number of generations in the growth cycle, and it can be valuable to normalize for this dependence, especially when comparing across environments [26]. However, defining the number of generations for a population with multiple genotypes growing at different rates is ambiguous. In the case of a single mutant competing with a wild-type, it is common to consider only the number of generations experienced by the wild-type strain, which is estimated by the log fold-change LFCwt = log Nwt(t)/Nwt(t − Δt). It is convenient to express the generations in base e to match the natural logarithm in the logit-encoding, but this could equivalently be done by converting all logarithms to base 2. Thus the relative fitness of a mutant per generation is defined by choosing Δt = LFCwt in Eq. (5). In the case of the logit-encoding, the relative fitness per-generation is where we have replaced ratios of relative abundances with ratios of absolute abundances in Eq. (6) and rear-ranged to express completely in terms of log fold-changes (Eq. (7)). The relative fitness per generation with the logit encoding is equivalent to the fitness statistic used in the LTEE [26] and other studies [73–77]. Some authors assume a fixed number of generations per growth cycle [12, 15], but since this may not be true, we use the term “per-generation” more strictly for fitness statistics where the wild-type generations are explicitly measured.

Relative fitness per-generation ranks mutants differently than relative fitness per-cycle

Even though relative fitness of a mutant per growth-cycle (Eq. (8)) and per generation (Eq. (9)) are both common statistics in microbial fitness measurements, they are not just equivalent quantities in different units. One major discrepancy between these fitness statistics occurs when the wild-type strain has no growth or experiences net death during the growth cycle because the number of wild-type generations is not well-defined (denominator of Eq. (9)) [5, 78]. This occurs for microbes under high drug concentrations, as well as for microbial populations under harsh environmental conditions as found in sediments or wastewater [79, 80]. Figure 1C shows four qualitatively different scenarios for a mutant competing with a wild-type strain, parameterized according to their LFCs. A comparison between relative fitness per-generation and per-cycle only makes sense when both the wild-type and the mutant show net growth (scenario 1 in Fig. 1C)

However, even when the LFC is positive, the relative fitness per-generation can produce different rankings compared to the relative fitness percycle. For example, Fig. 1D shows two mutant genotypes (red and grey) that have opposite rankings under these fitness statistics: the grey mutant has higher relative fitness per-cycle , but the red mutant has higher relative fitness per-generation . The disagreement in ranking requires positive covariation in the LFCs (red bowtie area in Fig. 1D), such that the mutant with higher LFC also induces a higher LFC in the wild-type (Sec. S5, Fig. S4).

Fitness statistics disagree over predicted ranking of single gene knockouts

While we have shown that relative fitnesses per-cycle and per-generation can lead to different mu-tant rankings in principle, we need to know whether this outcome occurs under realistic scenarios of microbial population dynamics. We thus simulate fitness rankings for gene knockout mutants using empirically measured growth traits, based on a previously published growth curve dataset for the single-gene knockout collection in Saccharomyces cerevisiae [29, 81]. We estimate growth traits from growth curves (Fig. 2A; Methods; Sec. S6) and find large variation in lag time, growth rate, and biomass yield for the single gene knockouts (grey dots in Fig. 2B,C; also see Fig. S5); despite the knockouts affecting multiple traits (pleiotropy), there are only weak correlations between them. To estimate relative fitness per-cycle (Eq. (8)) and per-generation (Eq. (9)), we simulate a pairwise competition for each knockout against the wild-type (Fig. 2D) using a consumer-resource model with a single limiting nutrient (Methods) [82, 83]. While the relative fitness per-cycle and per-generation are highly correlated across mutants over-all (Fig. S6), there are major differences in the ranks of individual mutants (Fig. 2E). For example, measuring the relative fitness per-generation ranks one beneficial mutant (grey dot highlighted with blue circle in Fig. 2E) 145 positions higher than quantifying relative fitness per-cycle (where higher rank corresponds to higher fitness). Indeed, when we plot the set of competitions in this fitness ranking on the LFC diagram, we find that there is many pairs of points that have positive covariation in the mutant and wild-type LFC and thus give rise to ranking differences (compare Figs. 2F and 1D) despite a negative covariation overall (compare Figs. 2F and S4B).

Comparison of mutant fitness rankings with different statistics on empirical trait variation.

(A) Overview of the growth curve dataset and the estimated growth traits for the knockout library of Saccharomyces cerevisiae (Methods). (B) Covariation between estimated steady-state growth rate g and lag time, λ across all mutant strains (grey dots; Pearson correlation coefficient r = −0.17, p = 7 × 10−30) as well as wild-type replicates (orange dots; r = − 0.16, p = 0.002). The reference wild-type strain for our pairwise co-culture simulations is defined by the median trait values (black cross) of all wild-type replicate. (C) Covariation between measured steady-state growth rate g and biomass yield Y across all mutant strains (grey dots; r = 0.21, p = 8 × 10-44) as well as wild-type replicates (orange dots; r = − 0.06, p = 0.25). (D) Overview of pairwise co-culture simulations. For each mutant strain (orange), we simulate a competition growth cycle against a reference wild-type strain using the estimated traits (panel A) and laboratory parameters for the initial condition (N0 = 0.05 OD, R0 = 111 mM glucose, x = 0.5; Methods) and quantify relative fitness of the mutant in different statistics (Eq. (8),Eq. (9)). (E) Rank disagreement between relative-fitness per-generation and per-cycle . For each fitness statistic, we calculate the mutant ranking (higher rank means higher fitness and mutants with equal fitness are assigned the lowest rank in the group). The rank difference is defined as the rank in minus the rank in . (F) Covariation between wild-type and mutant fold-change across all simulated competitions, with mutant strains (grey dots) and wild-type replicates (orange dots). For each wild-type replicate, we simulate a pairwise co-culture competition against the reference wild-type strain. We highlight the mutant with the greatest rank difference (blue halo) in panel E and F, and its corresponding bow-tie area of misranking (compare Fig. 1D).

What features of the population dynamics are responsible for the schism between relative fitness per-cycle and per-generation? We explore this by systematically varying the mutant trait distributions and initial conditions in our simulations (Fig. S7). This demonstrates that the fitness statistics and disagree on the rank-ing if the mutants have diminished biomass yield and are competed at a high initial frequency (50%) against the wild-type (rows A and D in Fig. S7). Intuitively, these conditions are necessary because they allow mutants to alter the wild-type LFC. Even though the wild-type growth rate is unaffected by mutants, the wild-type LFC will change in the presence of mutants that have lower yield because they reduce the time to resource depletion (row D in Fig. S7). A change in wild-type LFC also requires the mutant to be present at high relative abundance initially (row A in Fig. S7; see Sec. S7 for detailed conditions). By this mechanism, it is possible to concoct a set of growth traits where relative fitness per-generation and per-cycle deliver completely opposite rankings (Fig. S8).

As a further example, we test how the choice of fitness statistic affects conclusions from the LTEE. We reanalyze competition data from the LTEE [11] using the relative fitness per-generation (equivalent to the original definition used in those studies, ) and relative fitness per-cycle (Sec. S8). We see that the relative fitness statistics disagree on the rank-ing of some evolved lineages overall (Fig. S9) and at a given time point (Fig. S10). We confirm that the long-term fitness trend in the LTEE is the same for and (Fig. S11), but it is possible to construct hypo-thetical scenarios of evolution where this would not be the case (Fig. 3A-B). Beyond the qualitative mismatch between these fitness statistics, and can also lead to different conclusions for the presence of epistasis between multiple mutations (Fig. 3C-F,Fig. S12). For example, measuring relative fitness per-generation will detect negative magnitude epistasis between a mutation that only affects lag time and a mutation that only affects biomass yield (Fig. 3C), but measuring relative fitness per-cycle shows no epistasis (Fig. 3D). This is because mutations that increase biomass yield increase the wild-type LFC and, in this case, decrease the relative fitness per-generation (compare Fig. S12A and C). By the same mechanism, relative fitness per-cycle detects epistasis between mutations that affect growth rate and biomass yield (Fig. 3F), but no such epistasis is present in relative fitness per-generation (Fig. 3E). This again demonstrates that it is possible to arrive at different biological conclusions from the same experimental data de-pending on the choice of fitness statistic.

Potential consequences of the choice of fitness statistic for the interpretation of evolutionary data.

(A) Hypothetical scenario for the trait evolution in a long-term evolution experiment. An evolving population decreases in lag time (orange line), increases in growth rate to a maximum (blue line) and keeps decreasing in biomass yield (green line). This trend is similar to initial observations from the LTEE [94] (B) Corresponding long-term trend in relative fitness based on the trait evolution in panel A. We estimate relative fitness per-cycle (grey line) and per-generation (red line) every 250 generations. Dotted grey lines mark the end of trait evolution in lag time and growth rate in panel A. For the actual fitness trend in the LTEE, see Fig. S11. (C) Epistasis plot for lag time and yield using relative fitness per-generation . Colored dots show the fitness for a single mutant with shorter lag time (blue dot), a single mutant with higher biomass yield (red dot) and a double mutant with both mutations (purple dot). (D) Epistasis plot for lag time and yield in relative fitness per-cycle . (E) Epistasis plot for growth rate and yield in relative fitness per-generation . Colored dots show the fitness for a single mutant with higher growth rate (blue dot), a single mutant with higher biomass yield (red dot) and a double mutant with both mutations (purple dot). (F) Epistasis plot for growth rate and yield in relative fitness per-cycle . All epistais plots are based on 50:50 competition growth cycles with the wild-type (compare panels C,D and Fig. S12A-C, compare panels E,F and Fig. S12D-F).

Higher-order effects distort relative fitness measured in bulk competitions

So far we have focused on measuring relative fitness in pairwise competition, but this is usually not practical for large numbers of mutants. Measuring traits of genotypes in monocultures to predict their pairwise relative fitness (Sec. S1) is convenient, but when we test this strategy using our simulation framework we find that most traits perform poorly (Sec. S9, Fig. S13) (with some success for the AUC, but this success depends on the choice of time scale; Fig. S14). Another approach to measuring relative fitness is to use bulk competition experiments, where many mutant genotypes compete simultaneously in a single culture and each genotype is tracked through DNA barcode sequencing [39, 44]. However, this raises the question of how well relative fitness of a mutant in bulk competition corresponds to its relative fitness in a pairwise competition with the wild-type, since the growth in bulk might be influenced by the presence of other mutant genotypes, a phenomenon known as a higher-order interaction [83–85].

An important choice in bulk competition experiments is the relative abundance of the library of all mutant genotypes compared to the wild-type. The invasion of a spontaneous mutation into an existing population is best captured by pairwise competitions with low initial relative abundance of the mutant (Case I, Fig. 4A), and one way to recreate this scenario in a bulk competition is to use a low relative abundance for the mutant library overall (Case II, Fig. 4A) [40, 48–52]. A practical problem with Case II is that individual genotypes in the library will have low absolute abundances, which leads to stochasticity in the population dynamics and the sequencing preparation [13, 86, 87]. Therefore, it is common to compete the mutant library by itself (Case III, Fig. 4A) [13, 15, 17–19, 88].

The choice of library abundance and reference group in bulk competition experiments.

(A) Overview of a pairwise competition experiment (upper row) and multiple scenarios for bulk competition experiments (middle and bottom row) with different initial fraction of the mutant library (colored ovals) in the inoculum (open box). For each scenario, we show a schematic growth cycle (log absolute abundance) in the inset on the right. (B) Schematic relative abundance trajectories for a mutant compared to two alternative subpopulations. We distinguish between the total relative abundance xi with respect to the population as a whole (height of green band in the top box) and the pairwise relative abundance xiwt with respect to the wild-type (height of green band in the bottom box; Eq. (18). We indicate the sign of total relative fitness (Eq. (22)) and pairwise relative fitness (Eq. (23)) on the right. (C) The absolute error between bulk and pairwise competition experiments. The total relative fitness (grey dots; Eq. (22)) and the pairwise relative fitness (red dots; Eq. (23)) for empirical knockouts (Fig. 2A) in bulk competition growth cycle with low mutant library abundance (panel A, case II; Methods). The absolute error is defined as the bulk fitness statistic minus the relative fitness in pairwise competition (Eq. (S81)). In the inset, the absolute error for pairwise relative fitness (Eq. (23)) for a bulk competition growth cycle with high mutant library abundance (blue dots; case III). The x-axis and the red dots in the inset are identical to the main plot. (D) The relative error in bulk competition experiments as a function of mutant library abundance in the inoculum. Each line corresponds to a knockout in our dataset, and represents the relative error between the pairwise relative fitness in bulk competition and the relative fitness in pairwise competition (Eq. (S91)). In black lines, we show the recommended mutant library abundance for our dataset based on Eq. (10) (xlib ≈ 24.6%) and based on Eq. (S109) (xlib ≈ 0.02%, Sec. S15).

A second important choice is whether to quantify fitness of a mutant relative to the whole population (“total relative fitness”; Fig. 4B, top panel) [18, 19, 88] or relative to another specific genotype, like the wild-type (“pairwise relative fitness”; Fig. 4B, bottom panel; Methods) [40, 48–52]. In the case of a mutant library growing by itself (Case III, Fig. 4A), it is still possible to estimate a pairwise fitness by using known neutral mutants as the reference population [15, 17] and calculating fitness relative to this group (Secs. S10 and S11).

To test the consequences of these choices, we simulate a bulk competition (batch culture) using the trait data for yeast single-gene knockouts and compare the es-timates of total and pairwise relative fitness to fitness measured in pairwise competition, which we use as the ground truth (Methods). Figure 4C shows that the total relative fitness has systematically higher error than the pairwise relative fitness does, since the total fitness measures the genotype relative not just to the wild-type but to all other mutants as well (Sec. S12). For example, a mutant that grows identically to the wild-type (neutral phenotype) has zero increase in the pairwise relative abundance, but may have a net increase in the total relative abundance due to the poor growth of other mutants (compare top and bottom panel in Fig. 4B). This affects the total relative fitness of all mutants in a uniform way (Fig. 4C, Sec. S12), such that total and pairwise relative fitness completely agree in the ranking of genotypes (Fig. S15A).

While the pairwise relative fitness is the preferred method for estimating fitness in bulk competitions, the presence of higher-order interactions means it still deviates from the fitness in pairwise competitions. For the specific population dynamics in our simulation (Methods; Sec. S13,), we decompose the higher-order interaction into two terms, a fitness-independent term and a fitness-dependent term that acts as an amplifier for fitness values (Fig. S16; Sec. S14). Intuitively, the population in bulk competition consumes resources more slowly and gives more time for growth rate differences to accrue. The higher-order effects were tested by Levy et al. [13], who compared fitness estimates for ca. 30 beneficial lineages in bulk to the fitness estimates from pairwise competition and found that bulk fitness was typically higher, with some variation from lineage to lineage, consistent with our results here.

Can we reduce these deviations by changing the relative abundance of the mutant library (Fig. 4A)? When we repeat our simulation using a high mutant library abundance (Case III, Fig. 4A), we find that the absolute error for pairwise relative fitness increases proportionally (inset in Fig. 4C). Clearly, minimizing the abundance of the mutant library minimizes the strength of higher-order interactions (also reducing ranking disagreement (Fig. S15B,C).

As low mutant abundances create practical problems for experiments, it is valuable to identify a maximum abundance for the mutant library that keeps the error from bulk fitness estimates below a desired threshold. It is convenient to express this threshold in relative rather than absolute terms (compare Fig. S17 to Fig. 4C). Figure 4D shows the relative error from higher-order interactions across a range of mutant library abundances and using the specific population dynamics of our model, we derive the following rule: Assuming that the mutants in the library only have variation in growth rate, the mutant library abundance should be below where ϵ is the desired threshold on the relative error, gwt is the wild-type growth rate, and glib is the growth rate of the library as a whole (Sec. S15). In the case of our single-gene knockout library, Eq. (10) predicts a maximum library abundance of 24.6% based on the wild-type and library growth rate (gwt = 0.406, glib = 0.389) for a relative error ϵ = 1%. Figure 4D shows that this maximum library abundance keeps the relative error below 1% for high-fitness mutants (bright yellow), because they are dominated by growth rate effects, but fails for mutants close to neutrality because they have a trade-off between growth rate and lag time, which the estimate in Eq. (10) neglects. It is possible to derive a more precise bound that keeps the relative error below 1% for all mutant genotypes (here: xlib = 0.02%, see vertical line in Fig. 4D), but this requires prior knowledge of the trait covariation in the mutant library (Sec. S15).

Discussion

Best practices for quantifying mutant fitness in high-throughput experiments

In this work, we have introduced a conceptual framework (summarized in Fig. 5) to derive common statistics of relative fitness from three essential choices:

Flow-diagram for the quantification of relative fitness from time-series data.

Given the relative abundance of the mutant genotype at two consecutive timepoints (e.g. the start and end of a growth cycle), the user has to choose an encoding (Fig. 1A-B) and the time-scale for evaluating the change in relative abundance (Eq. (5)). In bulk competition experiments, multiple definitions of relative abundance are possible depending on the choice of the reference subpopulation (compare Fig. 4B; Methods). Each combination of these choices (dashed black lines) leads to a different fitness statistic and we summarize our recommendations (thick black line) in the discussion.

The choice of the state variable for relative abundance (encoding m; Eq. (2) and Fig. 1B).

The choice of the time scale for the change in relative abundance (Δt; Eq. (5)).

In bulk competition experiments, the choice of sub-population that acts as the reference for relative abundance of a chosen mutant genotype (Fig. 4B).

The combination of these choices leads to a range of fitness statistics, including those commonly used in population genetics, experimental evolution, and transposon-insertion screens (Fig. 5). However, as we compare these statistics in simulated competition experiments, we find that the choice of time scale can lead to different mutant rankings (Figs. 2D and 2E) and the choice of the reference subpopulation can lead to a systematic offset in fitness values (Fig. 4C). Based on these insights, we recommend the following choices for quantifying relative fitness of a mutant:

Use the logit encoding of relative abundance, because under the null model of logistic dynamics this linearizes the trajectory of relative abundance and regularizes measurement noise (Fig. S3).

Use a fixed extrinsic time scale (e.g., a single growth cycle), rather than an intrinsic time scale (e.g., the number of generations) which introduces an additional factor of variation between competitions.

In bulk competition experiments, use the pairwise fitness of the mutant relative to the wild-type reference subpopulation (either by including a barcoded wild-type or by grouping neutral mutants into a virtual wild-type) because this more closely matches the relative fitness in pairwise competition.

Since relative fitness measurements in bulk competitions inevitably carry an error from higher-order interactions, we recommend minimizing relative abundance of the mutant library as practically feasible (Fig. 4D). For a given error tolerance, Eq. (10) gives an estimate for the maximum library abundance.

If measuring direct competition is not feasible, we empirically find that the area under the growth curve (AUC) is the best approximation of the true fitness ranking (Fig. S13) but one must carefully choose the time scale (Fig. S14). Our recommendations agree with previous criticisms of relative fitness per-generation [5, 8, 9, 78] and total relative fitness [60, 89] but differ from the standard practice in high-throughput genetic screens [18, 19, 88] and many evolution experiments [33, 75], including the LTEE [11, 26]. The choice of the reference group is a well recognized issue in high-throughput evolutionary studies, in which one first estimates a total relative fitness and then subtracts a correction based on the fitness of the wild-type [12, 50, 52] or a mean population fitness [13, 41, 42, 51, 89]. In contrast, we recommend to choose the reference group at the level of relative abundance (Fig. 4B; Eq. (18))

Consequences of fitness quantification choices for microbial ecology and evolution

The quantitative differences between relative fitness statistics affects our ability to make evolutionary predictions. In particular, the fixation probability of a spontaneous mutation depends on the magnitude of relative fitness compared to other timescales like mutation and drift [57, 58], and in microbial populations with clonal interference, it depends on the entire distribution of fitness effects [59, 90]. For example, multiplying the mutation rate with an incompatible measure of fitness would lead to predicting the wrong speed of adaptation. In the context of multiple mutations in the same cell, different fitness statistics can lead to different conclusions about the presence of magnitude epistasis (Fig. 3C-F), and in the context of gene essentiality tests, they may affect the outcome of a significance test (e.g., using log or logit with the test statistic in Ref. [18]).

Using different fitness statistics can even lead to differences in mutant rankings (see Fig. S8 for an extreme case). Since measurements often serve as a first screen to narrow down the investigation to the top set of genes [17, 40], a difference in mutant ranking means the investigation might miss out on relevant genes because of the choice of fitness statistic.

Other sources of discrepancy in quantifying mutant fitness

Besides the conceptual choices of fitness quantification discussed in this article, experimental limitations can create discrepancies between replicate fitness measurements of the same mutant [36, 50, 91]. For example, relative abundance measurements entail sampling uncertainties (when we sample liquid for colony counting or DNA sequencing) [44, 60, 61], copy number variation [18, 44] as well as PCR jackpots and sequencing read errors [13, 86, 87]. Furthermore, there will inevitably be some variation in initial condition between replicates and fluctuations during the fitness assay [36, 87, 91]. On one hand, the conceptual choices for quantifying fitness can mitigate experimental uncertainty. For example, the logit encoding normalizes the sampling errors over the time series of relative abundance (Fig. S3B). On the other hand, there is sometimes a trade-off between conceptual choices and experimental precision, such as the choice of the initial mutant library abundance, which needs to be low to minimize higher-order interactions and yet high enough to minimize sampling uncertainty [44]. Future work will need to elucidate the interplay between conceptual and experimental sources of discrepancy in fitness measurements.

How general is the disagreement between fitness statistics?

The discrepancy between relative fitness per-cycle and per-generation (Figs. 1D and 2E), as well as the offset in the total relative fitness compared to pairwise fitness (Fig. 4C) will hold under any form of population dynamics. However, the details of the population dynamics do matter for the higher-order interactions that cause the discrepancy between relative fitness from bulk and pairwise competitions (Fig. 4D). The empirical evidence for the presence and mechanisms of higher-order interactions in microbial populations is limited [83–85]. Nevertheless, the fact that we see the fact that we see higher-order interactions even under competition for a single resource suggests that these effects will be present under more complex dynamics. Indeed, a comparison of mutant fitness measurements found systematic overestimates for fitness from bulk compared to pairwise competitions [13]. Since this could not be explained by the offset from using total relative fitness, it is indicative of higher-order interactions. Future work should provide more comparisons between relative fitness from bulk and pairwise competitions to better understand higher-order interactions.

Methods

Inferring growth traits for the single-gene knockout collection in Yeast

We use a previously published dataset [29] for the single-gene knockout collection in Saccharomyces cerevisiae [81], where the authors track growth of each genotype for 47 hours in monoculture, using microwell plates with defined growth medium. We download this growth curve dataset from the PROPHECY database [92], choosing specifically the dataset measured in Synthetic Defined medium. We last accessed the PROPHECY website (http://prophecy.lundberg.gu.se/) in March 30, 2020, but as of May 30, 2024, the website no longer seems accessible and included a reformatted version of the data with our code repository (https://github.com/justuswfink/24FitnessQuantification). From the raw timeseries of optical density, the authors subtracted a background correction based on blank wells, applied another correction for nonlinearity at high optical densities and smoothed the growth curve to remove electrical noise [29].

We start with the curves in the published dataset (9951 curves) and apply further trimming and smoothing steps (Sec. S6). From this data we calculate the time series of instantaneous per-capita growth rate N −1dN/dt (where N is the optical density) and identify time windows where the rate is approximately constant, which we interpret as distinct growth phases (Sec. S6). We only include curves that have a single phase of constant exponential growth followed by a stationary phase of approximately zero growth (9424 curves).

We quantify the biomass yield, maximum growth rate, and the lag time from each remaining curve as follows. First, we estimate the initial abundance Ninitial (average optical density over first three time points) and the final abundance Nfinal (average optical density over stationary phase) from the growth curve. Then we calculate the biomass yield as Y = (Nfinal − Ninitial)/R(0), where R(0) = 111 mM (20 g/L) is the initial concentration of glucose (assuming glucose is the single limiting resource). To estimate the maximum growth rate, we average the instantaneous growth rate over the exponential phase. Finally, we estimate the lag time from the intersection of the log initial abundance (log N0) with the slope of the maximum growth rate during the exponential growth phase [93]. We excluded curves with negative initial OD, curves with negative inferred lag times, and curves with a low quality of fit between the measured time series and a simulated curve based on the inferred trait values (R2 < 0.95; see below for the model). Our final database includes trait estimates for 9195 curves, which represents 92.4% of the original dataset [29].

Each single-gene deletion strain was measured in two technical replicate growth curves, using a second plate with identical layout in the same plate reader. Some genotypes have only one estimate of the growth traits because the other replicate did not pass our filters (273 genotypes), but for most genotypes we retain two replicate estimates in our final dataset (4163 genotypes); a few genotypes even have three (2 genotypes) or four replicate estimates (54 genotypes) because these genotypes were included multiple times by the original authors [29]. From these replicates we finally calculate the average yield, growth rate, and lag time for each geno-type. The dataset also contained many replicate growth curves of the wild-type strain, 374 of which passed our filters. Since wild-type traits inferred from these replicates had large variation (potentially due to measuring the wild-type across many different plates and days), we define the wild-type trait as the median, not the mean. Although the wild-type measurements included in this dataset have large trait variation (orange dots in Fig. 2B,C), the fact that the variation across knockouts is significantly greater for growth rate (Levene’s test p = 1.3 × 10−9) and lag time (Levene’s test p = 0.025) and that the replicate measurements of gene knockouts are correlated (Fig. S5) suggests that these trait measurements do capture true genetic variation. In contrast, the variation in biomass yield across knockouts is not significantly different from the variation across wild-type replicates (Levene’s test p = 0.44), suggesting that the variation in biomass yield is driven by non-genetic factors like the abundance or batch of resources used for each growth measurement.

Simulating population dynamics in competition experiments

Throughout this work, we model the competition between a set of genotypes during a single batch culture growth cycle using the following model of population dynamics [82, 83]: where Nj is the absolute abundance, xj(0) the initial relative abundance, gj(t, R) the growth rate, and Yj the biomass yield of genotype j, where the genotypes include a wild-type as well as one or more mutants. The initial absolute abundance of all genotypes together is N0, and the concentration of the single limiting resource is R with initial value R0. Note that Eq. (11b) assumes that cells consume resources only for biomass growth and not for maintenance of existing biomass.

To capture the sigmoidal shape of typical growth curves with a lag phase, exponential growth phase, and saturation phase (e.g., Fig. 2A), we model the dependence of growth rate on time and resource concentration as where Θ is the step function:

In this model, a genotype has an initial lag phase of time λj where no growth occurs, followed by a phase of constant exponential growth at rate gj, and ending when the resource concentration R reaches zero. The time to resource depletion, which we also call the saturation time, is defined by the implicit equation R(tsat) = 0. The simple form of the growth response (Eq. (12)) means that the absolute abundance of genotype j at saturation is and its log fold-change is

To calculate relative fitness per-generation (Eq. (9)) or per-cycle (Eq. (8) with Δt = 1), we numerically determine the saturation time tsat by integrating the biomass and resource dynamics (Eq. (11)) as described in previous work [82, 83]. We note that it also possible to derive an approximate expression for the saturation time tsat which we use for our theoretical calculations (Sec. S7).

Testing total and pairwise relative fitness in bulk competition experiments

We explore total and pairwise relative fitness as two alternative measures of fitness in bulk competition experiments, and defined as follows. Consider a population with multiple genotypes such that each genotype i has absolute abundance Ni(t) at time t. The total relative abundance of the genotype is

The “total” here refers to the fact that this is the abundance of genotype i relative to all other genotypes in the population. Similarly, we can think of the definition of relative fitness discussed in the main text (Eq. 3) as the total relative fitness of this genotype under an encoding m:

However, sometimes we want to track the dynamics of a genotype relative to another specific genotype; for example, to follow the fate of a mutant against the wild-type in a bulk competition experiment. For a pair of genotypes i ≠ j, we thus define the pairwise relative abundance which is the relative abundance of genotype i in a subpopulation of genotype i and j. To predict the change in the pairwise relative abundance, we can use the slope which defines the pairwise relative fitness of genotype i with respect to genotype j. In the special case of a population with only two strains, the pairwise relative fitness and total relative fitness are identical but they may differ with more than two genotypes (Sec. S12).

It is also possible to define the total and pairwise relative fitness statistics as finite differences over a time interval, rather than as instantaneous derivatives. For example, we can define them over a growth cycle starting at t = 0 and ending at t = tsat:

For these fitness statistics in the bulk competitions, we use the logit encoding where we have rewritten in the form that it is pre-sented in bulk competition experiments [15, 51, 52]. The logit encoding has mathematical advantages for coarse-graining the relative fitness of genotype groups (Sec. S10) but using the log encoding is another common choice in the literature [17, 41, 50]. However, the relative abundances of the individual mutants in our simulated bulk competition experiments are low enough that these two encodings are approximately equivalent (logit x ≈ log x for x ≪1).

Unlike the comparison between per-cycle and per-generation relative fitness where we focused on rank differences (Fig. 2E), here we can evaluate absolute differences in fitness estimates because the total and pairwise fitness in bulk fitness are measured in the same units as the relative fitness in pairwise competition. For all mutant genotypes, we calculate the absolute error between these bulk fitness estimates (Eq. (22),(23)) and the rel-ative fitness in pairwise competitions (Eq. (6)), which we take as the ground truth (Fig. 4C). Note that the pairwise competition could depend on the initial relative abundance of the mutant; we have chosen a very low rel-ative abundance (10−6) that mimics a mutant arising de novo and where this dependence is very weak.

eLife Assessment

This manuscript applies a theoretical analysis to two published datasets on yeast and bacterial evolution to compare different ways of quantifying fitness. It makes an important advance by clarifying how discrepancies can arise by using different approaches and provides recommendations for best practices. While the evidence is solid, some improvements in the presentation of the data and a greater focus on the causes of the discrepancies between the various fitness estimates would strengthen the paper further.

Reviewer #1 (Public review):

The authors point out that the fitness estimates obtained from different experimental assays (monoculture, pairwise competition, or bulk competition) are not generally equivalent, not even with regard to the fitness ranking of different genotypes. Using a computational model based on experimentally measured growth phenotypes for knockout strains in yeast, as well as data from Lenski's Long Term Evolution Experiment (LTEE), they derive a set of best practice rules aimed at extracting the optimal amount of information from such experiments.

The study is very complete on a technical level and I have no suggestions for further analyses. However, I feel the readability and the conceptual focus of the manuscript could be significantly improved by rearranging the material with regard to the contents of the main text vs. the Methods and the Supplement. Detailed recommendations:

(1) Regarding readability, the large number of references to material in the Methods and Supplement fragment the main text and make it difficult to follow.

(2) Conceptually, it seems to me that the current presentation obscures the reasons why we should care about fitness in the first place. In the first paragraph of Results, the authors define fitness "as any number that is sufficient to predict the genotype's relative abundance x(t) over a short-time horizon". To me, this seems like an extremely narrow and not very interesting definition. Instead, I view fitness as an intrinsic property of a genotype that allows us to predict its performance

under a range of conditions, including in particular conditions that are different from the experimental setup that was used to obtain the fitness estimates. The latter viewpoint is well expressed in Supplementary Section S1, where the authors discuss the notion of fitness potential. I would recommend to move at least part of this discussion to the main text. By comparison, the arguments in favor of the logit encoding that currently opens the Results session are rather straightforward and could be shortened significantly.

(3) Similarly, the modeling strategy used in this work is quite subtle and needs to be explained more fully in the main text. The authors use growth traits (lag time, growth rate, and yield) extracted from monoculture experiments on a yeast knockout collection and feed them into a specific mathematical model to simulate pairwise and bulk competition scenarios. Since a key claim of the work is that monoculture experiments are generally poor predictors of competitive fitness, the basis for this conclusion and the assumptions on which it is based need to be described clearly in the main text. In the current version of the manuscript, this information has

been largely relegated to the Methods section.

Reviewer #2 (Public review):

Summary:

The manuscript "Quantifying microbial fitness in high-throughput experiments" provides a comprehensive analysis of the various approaches to quantifying fitness in microbial evolution, focusing on three primary factors: encoding of relative abundance, time scale of measurement, and the choice of reference subpopulation. The authors systematically explore how these choices impact fitness statistics and provide recommendations aimed at standardizing practices in the field. This manuscript aims to highlight the impact of differing fitness definitions and the methodologies utilized for analysis and how that can significantly alter interpretations of mutant fitness, affecting evolutionary predictions and the overall understanding of genetic interactions in the experiments. Although this manuscript focuses on a critical issue in the quantification of fitness in high throughput experiments, it heavily relies on only one experimental dataset (Warringer et al 2003) and one organism i.e, Yeast (Saccharomyces cerevisiae) grown in a defined medium, the environmental influence is not completely captured. While the theoretical framework is strong, more experimental examples with more organisms (i.e., more datasets) in their analysis and comparison would enhance the manuscript, especially its conclusion.

Strengths:

The choices for quantifying fitness in evolution experiments are critical and highly relevant given the increasing prevalence of high-throughput experiments in evolutionary biology. The authors methodically categorize fitness statistics and their implications, providing clarity on a complex subject. This structured approach aids in understanding the nuances of fitness measurement. The manuscript effectively highlights how different choices in fitness measurement can influence fitness rankings and the understanding of epistasis, which is important for modeling evolutionary dynamics.

Weaknesses:

The theoretical framework is robust, but the manuscript could benefit from more empirical examples to illustrate how different fitness quantification methods lead to varied conclusions in experiments. The discussion on the choice of reference subpopulation could be expanded with the influence of the environment or the condition. Different types of reference groups might yield different implications for fitness calculations, and further elaboration would enhance this section. The authors overgeneralize some findings; for instance, the implications of fitness measurement choices could vary significantly across different microbes or experimental conditions. A more detailed discussion would strengthen the conclusion.

Overall, this manuscript is a significant contribution to the field of evolutionary biology, addressing a critical issue in the quantification of fitness but lacks more experimental support to make it a wider claim. By systematically exploring the factors that influence fitness measurements, the authors provide valuable insights that can guide future research - the framework is computationally thorough but needs a more detailed explanation of concepts instead of generalizing. Further work is needed, particularly to incorporate empirical examples and expand certain discussions to include environmental variation and their impact, which would improve clarity and applicability.

Reviewer #3 (Public review):

Summary:

The authors present analyses of different fitness measures derived from empirical data from yeast knock-out mutants and the long-term evolution experiment (LTEE) with Escherichia coli to explore discrepancies and identify preferred methods to estimate relative fitness in high-throughput experiments. Their work has three components. They first discuss the different "encodings" of relative abundance data and conclude that logit transformations are preferred because they transform nonlinear abundance trajectories into linear trajectories with greater predictive power. Next, they compare per-generation with per-growth cycle relative fitness estimates inferred from simulations of pairwise competitions based on published growth traits for the yeast strains and on published pairwise competition measurements for the LTEE data. Both data sets show quantitative and qualitative (i.e. rank order) discrepancies of estimates across different time scales, which are highlighted by considering possible underlying causes (i.e. trade-offs between growth traits) and consequences (i.e. epistasis among mutations affecting different growth traits). Finally, the authors compare simulated pairwise and bulk (i.e. where many mutants compete during a growth cycle in a single environment) competition assays based on the yeast knock-out mutants and demonstrate an optimal ratio of collective mutants to wild-type strains that minimizes both sampling error and overestimation of fitness estimates when compared with pairwise competitions.

Strengths:

The study deals with a highly relevant topic. Fitness is central to general evolutionary theory, but also poorly defined and implies different traits for different organisms and conditions. For microbes, which are often used in evolution experiments, high-throughput experiments may yield different measures to quantify abundance over time, from individual growth traits to bulk competition experiments. Hence, it is relevant to consider discrepancies among those measures and identify preferred measures with respect to predicting population dynamics and evolutionary processes. The present study contributes to this aim by (i) making readers aware of differences among commonly used fitness estimates, (ii) showing that simulated (yeast) and calculated (E. coli) competitive fitness may differ across time scales, and (iii) showing that bulk competitions may yield relative fitness estimates that are systematically higher than pairwise competitions. The study is rather thorough on the theory side, with extensive derivations and analyses of various fitness measures using their resource competition model in the Supplementary Information. The study ends with a few practical recommendations for preferred methods to infer relative fitness estimates, that may be useful for experimentalists and stimulate further investigations.

Weaknesses:

The study has several limitations. Perhaps the most apparent limitation is the lack of a clear answer to the question of which fitness measure is best "in the light of first principles". The authors show clear discrepancies between fitness estimates across different time scales or using different reference genotypes in bulk competition and provide useful recommendations based on practical considerations (e.g. using pairwise competitions as the "golden standard"), but it remains unclear whether these measures provide the greatest value for the questions researchers may want to answer with them (e.g. predict shifts in genotype frequencies).

A second limitation is that the authors analyse fitness differences arising solely from resource competition, whereas microbes often interact via other mechanisms, e.g. the production of anticompetitior toxins, cross-feeding of metabolites, or lack of growth to enhance their persistence in stress conditions. Without simulations of these processes, understanding discrepancies among fitness measures is necessarily limited. In addition, the analysis of trade-offs between growth traits causing these discrepancies during resource competition seems confounded by biases in measurement error or parameter estimation, at least for growth rate and lag time (Figure 2B), where the replicate estimates for the wildtype show a similar negative correlation.

Third, the study does not validate relative fitness predictions from growth traits (as is done for the yeast mutants) with measured relative fitness estimates using competition assays, while such data are available, e.g. for the LTEE. This would strengthen their inferences about preferred fitness measures.

Fourth, the analysis of epistasis between mutations affecting different growth traits (shown in Figure 3) based on the LTEE data could be better introduced and analysed more comprehensively. Now, the examples given in panels C-F seem rather idiosyncratic and readers may wonder how general these consequences of using fitness estimates based on different time scales are.

Finally, the study is generally less accessible to experimentalists due to the extensive and principled treatment of specific population dynamic models and fitness inferences. This may distract from the overarching aim to identify fitness measures that are most accurate and useful for predictions of population dynamics and evolutionary processes. In this light, the motivation for the initial discussion of the importance of how to best encode relative abundance (Figure 1) is unclear. Also, the conclusion, that logit encoding is preferred, because it linearizes logistic growth dynamics and "improves the quality of predictions", is not further motivated. Experimentalists using non-linear models to infer fitness from growth curves or competition assays may miss the relevance of this discussion.

Author response:

The following is the authors’ response to the original reviews.

We thank both editors and the three reviewers for their constructive criticism of our work. As a result of these comments, we have made several significant revisions to the paper that we believe strengthen and clarify our major results:

(1) Following suggestions from Reviewers #1 and #3, we have have improved our introduction to the different fitness concepts (lines 105–148) and streamlined the discussion of the logit encoding (lines 175–190). In particular, we have moved the most technical points to the SI (Sec. S3).

(2) Based on criticisms of our usage of the population dynamics model from Reviewers #1 and #3, we significantly revised our explanation of the motivation and interpretation of this model (lines 284–310 and 323–336) and our discussion of the generalizability of these results (lines 678–728), including the possible effects of interactions besides resource competition.

(3) Following a request from Reviewer #3, we have expanded our analysis of epistasis to systematically test all possible double mutants between qualitative types of trait perturbations in the model. We have added a new main text figure (Fig. 3), new SI figures (Figs. S9–S15), a new subsection in the Results (lines 344–395), and corresponding new sections in the Methods (lines 864–892) and SI (Sec. S8).

(4) Following concerns from Reviewers #2 and #3 about the limited empirical data, we have expanded our analysis of the LTEE data (new main text Fig. 4, revised text on lines 416–439, and revised SI Figs. S16–S18) and have analyzed two new benchmarking datasets for bulk fitness to test our predictions (new main text Fig. 6, new Results subsection on lines 561–590, and new SI Figs. S24 and S25).

(5) Following the criticism of Reviewer #3 about the lack of a clear recommendation on fitness quantification that provides the greatest value for a given scientific question, we have better explained what we think the scientific consequences of fitness are as a motivation for our analysis (lines 82–88, 319–322, and 615–630) and replaced the final flowchart figure with a step-by-step guide in the Methods to implement our recommendations in practice (lines 964–982).

Reviewer #1 (Public review):

The authors point out that the fitness estimates obtained from different experimental assays (monoculture, pairwise competition, or bulk competition) are not generally equivalent, not even with regard to the fitness ranking of different genotypes. Using a computational model based on experimentally measured growth phenotypes for knockout strains in yeast, as well as data from Lenski’s Long Term Evolution Experiment (LTEE), they derive a set of best practice rules aimed at extracting the optimal amount of information from such experiments.

The study is very complete on a technical level and I have no suggestions for further analyses. However, I feel the readability and the conceptual focus of the manuscript could be significantly improved by rearranging the material with regard to the contents of the main text vs. the Methods and the Supplement. Detailed recommendations:

(1) Regarding readability, the large number of references to material in the Methods and Supplement fragment the main text and make it difficult to follow.

We understand the challenges these references pose to the flow of the main text; we have attempted to keep those references to a minimum, while ensuring that technical details of the work are fully documented and referenced for completeness.

(2) Conceptually, it seems to me that the current presentation obscures the reasons why we should care about fitness in the first place. In the first paragraph of Results, the authors define fitness “as any number that is sufficient to predict the genotype’s relative abundance x(t) over a short-time horizon”. To me, this seems like an extremely narrow and not very interesting definition. Instead, I view fitness as an intrinsic property of a genotype that allows us to predict its performance under a range of conditions, including in particular conditions that are different from the experimental setup that was used to obtain the fitness estimates. The latter viewpoint is well expressed in Supplementary Section S1, where the authors discuss the notion of fitness potential. I would recommend to move at least part of this discussion to the main text.

We appreciate the reviewer’s viewpoint and have moved that conceptual discussion from the SI to the beginning of the Results section to give readers a broader perspective on fitness (lines 105–148). We use “potential” in analogy with potential energy in physics and have clarified this on lines 126–135.

What we call fitness potential, like the other notions of fitness we discuss in this paper (relative and absolute fitness), is still specific to an environmental condition. Fitness as a property intrinsic to a genotype and independent of any environment, as the reviewer mentions, is an interesting concept but beyond the scope of this paper, which is focused on analyzing fitness measurements that are inevitably environment-specific and we have clarified this on lines 142–148. While it is true that this definition of fitness is narrow, it is what can be empirically measured directly, and thus we believe it is crucial to understand how to best interpret that data.

By comparison, the arguments in favor of the logit encoding that currently opens the Results session are rather straightforward and could be shortened significantly.

We agree and have condensed this section (lines 175–192).

(3) Similarly, the modeling strategy used in this work is quite subtle and needs to be explained more fully in the main text. The authors use growth traits (lag time, growth rate, and yield) extracted from monoculture experiments on a yeast knockout collection and feed them into a specific mathematical model to simulate pairwise and bulk competition scenarios. Since a key claim of the work is that monoculture experiments are generally poor predictors of competitive fitness, the basis for this conclusion and the assumptions on which it is based need to be described clearly in the main text. In the current version of the manuscript, this information has been largely relegated to the Methods section.

We agree that our motivation for the population dynamics model and growth curve data was not clearly explained. We have significantly revised this section of the Results in the main text (lines 284–310).

In particular, we recognize the potential for misunderstanding this material we do not intend the relative fitness values calculated from this model to be interpreted as predictions of the true relative fitness between yeast deletion strains. Rather, we use the population dynamics model for our proof of principle: that the most basic features of microbial population dynamics in laboratory experiments, as captured by this model (resource competition, lag phase, growth phase, saturation), are sufficient to create discrepancies between common fitness statistics used in these experiments (different encodings, time scales, choices of reference subpopulations). We have added a statement to highlight existing work on monoculture predictors for competition outcomes [32, 34, 36, 37] on lines 453–459.

Reviewer #1 (Recommendations for the authors):

In the discussion of the LTEE in Section S8, the authors write on page 8 that “we couldn’t fit the fitted values a,b in ref. 29 so we were unable to check it”. I don’t understand this sentence - is the claim that the fit in ref. 29 was incorrect?

We have clarified this point in the SI (now Sec. S9). Our point was not that the fit in Wiser et al. 2013 is incorrect, but merely that we could not find the exact values of the fitted parameters they obtained documented in their paper, so we could not compare our own fitted parameters directly to theirs.

Also, at the end of the section, the authors refer to theory work on the long-term fitness trend in the LTEE. Here, two early references arguing for a logarithmic increase in fitness could be mentioned as well:

International Journal of Modern Physics B 12,:361-391 (1998) Evolution and Extinction Dynamics in Rugged Fitness Landscapes Paolo Sibani, Michael Brandt, and Preben Alstrøm

J. Stat. Mech. (2008) P04014 Evolution in random fitness landscapes: the infinite sites model Su-Chan Park and Joachim Krug

We thank the reviewer for providing these two references and have added them to the list of previous works on long-term fitness trends at the end of the section (now Sec. S9).

Reviewer #2 (Public review):

Summary:

The manuscript “Quantifying microbial fitness in high-throughput experiments” provides a comprehensive analysis of the various approaches to quantifying fitness in microbial evolution, focusing on three primary factors: encoding of relative abundance, time scale of measurement, and the choice of reference subpopulation. The authors systematically explore how these choices impact fitness statistics and provide recommendations aimed at standardizing practices in the field. This manuscript aims to highlight the impact of differing fitness definitions and the methodologies utilized for analysis and how that can significantly alter interpretations of mutant fitness, affecting evolutionary predictions and the overall understanding of genetic interactions in the experiments. Although this manuscript focuses on a critical issue in the quantification of fitness in high throughput experiments, it heavily relies on only one experimental dataset (Warringer et al 2003) and one organism i.e, Yeast (Saccharomyces cerevisiae) grown in a defined medium, the environmental influence is not completely captured. While the theoretical framework is strong, more experimental examples with more organisms (i.e., more datasets) in their analysis and comparison would enhance the manuscript, especially its conclusion.

We have expanded our analysis of competition data from the Long-Term Evolution Experiment in E. coli (lines 416– 439), including adding a main text figure (Fig. 4) along with the three SI figures (Figs. S16–S18). We have also added two completely different data sets that directly test our predicted discrepancies in fitness estimates from bulk competition experiments. From this data we have added a new main text figure (Fig. 6), two new SI figures (Figs. S24 and S25), and a new section at the end of the Results (lines 563–590).

We wish to clarify, though, that the aim of this study is to develop theory on fitness quantification choices and minimal examples to demonstrate the potential for discrepancies between these choices. While we appreciate the reviewer’s interest in understanding how discrepancies in fitness statistics vary across organisms and environments, that is an empirical question beyond the scope of this paper.

Strengths:

The choices for quantifying fitness in evolution experiments are critical and highly relevant given the increasing prevalence of high-throughput experiments in evolutionary biology. The authors methodically categorize fitness statistics and their implications, providing clarity on a complex subject. This structured approach aids in understanding the nuances of fitness measurement. The manuscript effectively highlights how different choices in fitness measurement can influence fitness rankings and the understanding of epistasis, which is important for modeling evolutionary dynamics.

Weaknesses:

The theoretical framework is robust, but the manuscript could benefit from more empirical examples to illustrate how different fitness quantification methods lead to varied conclusions in experiments.

Please see our response to the previous comment on this point.

The discussion on the choice of reference subpopulation could be expanded with the influence of the environment or the condition. Different types of reference groups might yield different implications for fitness calculations, and further elaboration would enhance this section.

While we agree that studying how environmental conditions affect fitness is an important and interesting problem, it goes beyond the scope of this paper, which focuses on the basic theory of quantifying microbial fitness from highthroughput experiments. Applications of this theory to empirical questions about environmental variation would be best served by their own studies. We have added a statement clarifying this goal (lines 144–148).

We are unsure how the choice of reference subpopulation is related to this issue. In our view, if the goal of a mutant fitness measurement is to predict how that mutant would behave when arising spontaneously and competing against its immediate ancestor, the gold-standard reference subpopulation must always be the mutant’s immmediate ancestor, or another mutant that is known to be phenotypically equivalent to the ancestor (e.g., neutral mutants in the case of a large mutant library). Other choices of reference subpopulations would not provide directly meaningful information in this regard.

The authors overgeneralize some findings; for instance, the implications of fitness measurement choices could vary significantly across different microbes or experimental conditions. A more detailed discussion would strengthen the conclusion.

We certainly agree that the consequences of fitness quantification choices could vary significantly across organisms and environments; our goal for this paper is to demonstrate what discrepancies are possible in principle and in particular how they depend on basic features of microbial population dynamics (e.g., variation in yield). We have added two separate paragraphs in the Discussion section to address the generalizability of our results in the context of pairwise (lines 678–710) and bulk fitness measurements (lines 711–728).

Overall, this manuscript is a significant contribution to the field of evolutionary biology, addressing a critical issue in the quantification of fitness but lacks more experimental support to make it a wider claim. By systematically exploring the factors that influence fitness measurements, the authors provide valuable insights that can guide future research - the framework is computationally thorough but needs a more detailed explanation of concepts instead of generalizing.

We have improved our explanation of several of the important concepts. In particular, we have significantly revised our explanation of the population dynamics model (lines 284–310) to emphasize its role as a null model to demonstrate how fundamental aspects of microbial growth are sufficient to cause discrepancies between fitness statistics. We have also revised two paragraphs on the generalizability of our results in the Discussion section (lines 678–728).

Further work is needed, particularly to incorporate empirical examples and expand certain discussions to include environmental variation and their impact, which would improve clarity and applicability.

We have added a sentence at the beginning of the Results section to acknowledge the environmental dependence of fitness (lines 142–148). We believe further discussion of that issue is beyond the scope of this paper, as it would require a significant amount of additional data and/or environmental modeling.

Reviewer #2 (Recommendations for the authors):

In addition to the comments from the previous sections, other specific comments:

(1) Figure 5 needs to be populated with additional parameter details. For example, include brief descriptions of each parameter involved in the encoding, time scale, and reference choices. This will help users understand the implications of each choice. Adding these details will make the flow diagram more comprehensive, aiding researchers in implementing these steps more clearly.

Following this comment and another comment about this figure from Reviewer #3, we decided to replace this figure with a new Methods section with step-by-step instructions (lines 964–982).

(2) Duplication in Line 620: “Nevertheless, the fact that we see the fact that we see...” This redundancy needs to be corrected.

We thank the reviewer for pointing this out; we have rewritten this paragraph.

(3) More experimental data comparisons and their assessment concerning various microbial systems and multiple environmental conditions are recommended to support the claim.

Please see our responses to the related public comments.

Reviewer #3 (Public review):

Summary:

The authors present analyses of different fitness measures derived from empirical data from yeast knockout mutants and the long-term evolution experiment (LTEE) with Escherichia coli to explore discrepancies and identify preferred methods to estimate relative fitness in high-throughput experiments. Their work has three components. They first discuss the different “encodings” of relative abundance data and conclude that logit transformations are preferred because they transform nonlinear abundance trajectories into linear trajectories with greater predictive power. Next, they compare per-generation with per-growth cycle relative fitness estimates inferred from simulations of pairwise competitions based on published growth traits for the yeast strains and on published pairwise competition measurements for the LTEE data. Both data sets show quantitative and qualitative (i.e. rank order) discrepancies of estimates across different time scales, which are highlighted by considering possible underlying causes (i.e. trade-offs between growth traits) and consequences (i.e. epistasis among mutations affecting different growth traits). Finally, the authors compare simulated pairwise and bulk (i.e. where many mutants compete during a growth cycle in a single environment) competition assays based on the yeast knock-out mutants and demonstrate an optimal ratio of collective mutants to wild-type strains that minimizes both sampling error and overestimation of fitness estimates when compared with pairwise competitions.

Strengths:

The study deals with a highly relevant topic. Fitness is central to general evolutionary theory, but also poorly defined and implies different traits for different organisms and conditions. For microbes, which are often used in evolution experiments, high-throughput experiments may yield different measures to quantify abundance over time, from individual growth traits to bulk competition experiments. Hence, it is relevant to consider discrepancies among those measures and identify preferred measures with respect to predicting population dynamics and evolutionary processes. The present study contributes to this aim by (i) making readers aware of differences among commonly used fitness estimates, (ii) showing that simulated (yeast) and calculated (E. coli) competitive fitness may differ across time scales, and (iii) showing that bulk competitions may yield relative fitness estimates that are systematically higher than pairwise competitions. The study is rather thorough on the theory side, with extensive derivations and analyses of various fitness measures using their resource competition model in the Supplementary Information. The study ends with a few practical recommendations for preferred methods to infer relative fitness estimates, that may be useful for experimentalists and stimulate further investigations.

Weaknesses:

The study has several limitations. Perhaps the most apparent limitation is the lack of a clear answer to the question of which fitness measure is best “in the light of first principles”. The authors show clear discrepancies between fitness estimates across different time scales or using different reference genotypes in bulk competition and provide useful recommendations based on practical considerations (e.g. using pairwise competitions as the “golden standard”), but it remains unclear whether these measures provide the greatest value for the questions researchers may want to answer with them (e.g. predict shifts in genotype frequencies).

We agree on the importance of considering the scientific questions researchers want to answer in determining the best way to quantify fitness. We have revised both the Introduction (lines 82–88) and the Discussion (lines 615–630) to more clearly explain possible downstream questions researchers may wish to answer with fitness data, and thus why discrepancies in that data based on analysis choices may be important.

We believe that the text does provide a specific recommendation (second subsection of the Discussion, lines 635– 658) for how to quantify relative fitness: using the logit encoding (rather than other encodings), measuring fitness per-cycle (rather than per-generation), and using the wild-type or a phenotypically-equivalent proxy as reference subpopulation to calculate pairwise fitness in a bulk competition (rather than using the mutant library as a whole). This recommendation is based on first principles: the logit encoding is based on the principle of the logistic equation as the null model of relative abundance dynamics (lines 635–637), the choice of the per-cycle timescale is based on the principle that in non-steady state environments the time scale for measuring selection should not depend on the wild-type growth (lines 640–645), and the choice of reference population is based on the principle that a mutant’s fitness should serve as a predictor of its dynamics when arising de novo at low frequency and competing against its wild-type (lines 648–653).

A second limitation is that the authors analyse fitness differences arising solely from resource competition, whereas microbes often interact via other mechanisms, e.g. the production of anticompetitor toxins, cross-feeding of metabolites, or lack of growth to enhance their persistence in stress conditions. Without simulations of these processes, understanding discrepancies among fitness measures is necessarily limited.

We agree that other interactions are important in many microbial ecosystems and could affect measurements of fitness. We discuss the possibility of these other interactions and their potential consequences for fitness on lines 697– 710.

We focus on resource competition in this paper, however, for two reasons. One is that we are using it as a null model: resource competition is always present, and thus it provides an important baseline for discrepancies in fitness statistics in the absence of any other assumptions. Indeed, our results are that this minimal assumption alone is sufficient to produce a wide range of significant discrepancies, which provides the proof of principle that choices of fitness quantification matter. We have clarified this in a revised explanation of the population dynamics model on lines 294–304.

The second reason is that fitness measurements of the type discussed in this paper are typically performed on mutants that have only small genetic differences with their ancestor (e.g., a point mutation or gene deletion). While more complex interactions between such similar genotypes are not impossible, we expect them to be rare, in which case resource competition is the only interaction. Explicit modeling of other interactions is an important question for future work, but would require more detailed models and data of those phenomena, and thus would go beyond the scope of the present study. We have added a sentence to explain our emphasis on resource competition on lines 298–301 and 690–697.

In addition, the analysis of trade-offs between growth traits causing these discrepancies during resource competition seems confounded by biases in measurement error or parameter estimation, at least for growth rate and lag time (Figure 2B), where the replicate estimates for the wildtype show a similar negative correlation.

The tradeoff between growth traits was only an incidental observation and is not necessary for the fitness statistic discrepancies we analyze in this paper; the only important pattern in the growth traits is the existence of mutants with reduced yields (so as to reduce the wild-type log fold-change in a competition) as well as variation in one other trait under selection (lag time or growth rate in this model). We have clarified this mechanism on lines 328–336, which is demonstrated by Fig. S7. Since these tradeoffs are not relevant to the results and we agree that their significance may be unreliable due to the noisiness of the data, we have removed mention of them.

Third, the study does not validate relative fitness predictions from growth traits (as is done for the yeast mutants) with measured relative fitness estimates using competition assays, while such data are available, e.g. for the LTEE. This would strengthen their inferences about preferred fitness measures.

The goal of our modeling with the yeast growth trait data is not to test the ability to predict competition experiments from monoculture data; that has been the focus of previous studies [32, 34, 36, 37]. Rather, we use the population dynamics model for a proof of principle: that the most basic features of microbial population dynamics in laboratory experiments, as captured by this model (resource competition, lag phase, growth phase, saturation), are sufficient to create discrepancies between common fitness statistics used in these experiments (different encodings, time scales, choices of reference subpopulations). The yeast growth curve data merely provides realistic parameters for this model, to ensure we are studying a biologically relevant regime of the dynamics. To avoid this misconception, we have revised our explanation of this model and the data on lines 284–310.

Fourth, the analysis of epistasis between mutations affecting different growth traits (shown in Figure 3) based on the LTEE data could be better introduced and analysed more comprehensively. Now, the examples given in panels C-F seem rather idiosyncratic and readers may wonder how general these consequences of using fitness estimates based on different time scales are.

We agree that this analysis was incomplete and missed an opportunity to emphasize this important consequence of fitness quantification. We have thus expanded this analysis into a systematic test of all possible double mutants between qualitative types of trait perturbations in the model. We have added a new main text figure (Fig. 3), new SI figures (Figs. S9–S15), a new subsection in the Results (lines 346–395), and corresponding new sections in the Methods (lines 864–892) and SI (Sec. S8).

Finally, the study is generally less accessible to experimentalists due to the extensive and principled treatment of specific population dynamic models and fitness inferences. This may distract from the overarching aim to identify fitness measures that are most accurate and useful for predictions of population dynamics and evolutionary processes.

We appreciate this concern as we do hope to make the paper as broadly accessible as possible, especially to experimentalists who measure microbial fitness. To this end, we have reduced the technical discussion of encodings in the first section of the Results (lines 164–187); revised explanations of the population dynamics model (lines 284–310), importance of growth trait variation (lines 328–336), and epistasis (lines 346–395) to better emphasize the conceptual intuition of these parts; and added a step-by-step guide for our recommended best practices of quantifying fitness in bulk competition experiments (lines 964–982).

In this light, the motivation for the initial discussion of the importance of how to best encode relative abundance (Figure 1) is unclear. Also, the conclusion, that logit encoding is preferred, because it linearizes logistic growth dynamics and “improves the quality of predictions”, is not further motivated. Experimentalists using non-linear models to infer fitness from growth curves or competition assays may miss the relevance of this discussion.

The motivation for the discussion of encodings is that it is one of the choices made differently by researchers, mainly using either the logit (more common in experimental evolution and population genetics studies) or log encoding (more common in TnSeq analyses). As such we believe it is important to explain where this choice comes from (a transformation of relative abundance data to make it approximately linear in time, and thus amenable to characterization by a single slope parameter) and why we believe the logit encoding is more logical in most cases. We have streamlined and revised this subsection to make it clearer (lines 164–187).

Our argument for favoring the logit encoding in most cases is based on the logistic model being a null model for relative abundance dynamics (Sec. S3). In light of the reviewer’s comments, we have realized this may be confusing because there are two common usages of logistic dynamics that are biologically distinct. What we mean by logistic model is the dynamics of relative abundance x of a mutant in competition with other genotypes:

Here s turns out to be the relative fitness under the logit encoding. On the other hand, researchers also use a logistic ODE to describe the dynamics of absolute abundance N of a single strain in monoculture (e.g., as in a growth curve):

We believe the reviewer’s last point refers to Eq. (2), whereas our argument about the logit encoding is based on Eq. (1). We have added a note to clarify this distinction for the reader (lines 192–196).

Reviewer #3 (Recommendations for the authors):

In addition to my general comments in the public review, I have several more specific recommendations:

(1) Line 183-189: unclear why logit-based relative fitness is preferred. Abundance data are not typically binomial.

We agree this claim about abundance data was incorrect and have removed it. We have revised the section to focus on motivating the logit encoding from logistic dynamics of relative abundance as a null model for most systems (main text lines 175–187 and Sec. S3).

(2) Line 205: it may be mentioned that s(logit) is the same as the “selection rate constant” often used in microbial studies.

We have added a sentence clarifying the equivalence of the logit-encoded relative fitness to the selection coefficient in population genetics (lines 188–190).

(3) Line 368: why do mutations that increase biomass yield also increase WT LFC? Is this, because they grow slower and hence allow the WT more time to grow?

Mutants with higher yield allow the wild-type to achieve higher log fold-change because those mutants consume fewer resources per cell, which frees up more resources for the wild-type to consume and increase its overall growth. It’s not about growth rate or time, as this would occur even for mutants whose growth rates are identical to the wild-type’s. We have revised our explanation of how variation in growth traits differentially affects fitness statistics (lines 323–340) and epistasis (lines 361–378).

(4) Line 382-386: you may want to cite Ram et al. (2019, 10.1073/pnas.1902217116), who also did such analyses for experimental data from E. coli.

We have cited this work as Ref. [34].

(5) Line 415: perhaps use “bulk relative fitness” instead of “total relative fitness”, to contrast with “pairwise relative fitness”.

We acknowledge the language in this section can be subtle. However, “bulk” is not a sufficient identifier for the concept of total relative fitness as bulk competition experiments (with many genotypes competing simultaneously) can be used to measure either total relative fitness or pairwise relative fitness. (In pairwise competition experiments with only two genotypes, these two types of fitness are identical.) As such we adhere to our original language but have added words to clarify which type of experiment (bulk or pairwise) we are talking about in a given context (e.g., on lines 495–504).

(6) Line 451-453: why does a population in bulk competition consume resources more slowly than in pairwise competitions?

Mutant libraries used in bulk competition experiments usually include a large number of deleterious mutants, which grow more slowly than the wild-type. Thus these populations typically consume resources more slowly than a population in a pairwise competition would, where a large part of the population is the wild-type.

(7) Line 565: I don’t understand how one can compare relative fitness to other timescales.

Relative fitness, as we’ve defined it, has units of rate, since it describes the rate of change of relative abundance (or an encoding of it) over some time scale (e.g., a batch growth cycle or a generation). Therefore it can be compared to other times scales of the system, such the rate of new mutations arising or the rate of genetic drift fluctuations, as long as they are measured in the same units. This comparison is important to population genetics analyses, such as determining whether the population is in the strong selection-weak mutation limit or the clonal interference regime.

(8) Line 620 repeats text.

Thank you, we have revised this paragraph and removed the typo.

(9) Figure 1C+D: the link between the scenarios on the left and the graphs on the right may be better explained. For example, it may help to make explicit that the 4 scenarios in panel C show the same relative fitness per cycle and that mutant and wildtype have the same growth rate, but different growth periods in both scenarios in panel D. It is also unclear whether the grey dot links to the upper scenario in D.

We have clarified this issue in the caption and changed the colors to avoid this confusion.

(10) Figure 2E: it is unclear why “mutants with equal fitness are assigned the lowest rank”.

This was a technical comment about how to handle ties in our analysis of mutant rankings, but it is moot since no exact ties actually occur in our simulations. We have removed this remark to avoid confusion.

(11) Figure 2F: the axis labels are confusing, as for the WT estimates no LFC mutant exists. It would also help to make explicit in the legend against which WT replicate/reference strain each strain has competed.

We agree the inclusion of wild-type replicates in this plot was confusing and unnecessary, so we have removed them. The mutants compete against a wild-type with traits defined by their median values across all wild-type replicates; this is noted in Fig. 2A and the Methods section on our analysis of this data (lines 809–813).

(12) Figure 5: I am not sure this is needed, as its information is rather limited.

We agree and have removed this figure.

Version 2

Abstract

Few concepts are as central to evolution as is fitness, and yet the quantification of fitness is often ambiguous. In particular, high-throughput experiments to measure mutant fitness in microbes are increasingly common but vary widely in their definitions of fitness, which makes their results difficult to compare. What are the consequences of these different fitness statistics, and is there a best way to quantify fitness in a given context? Here we systematize the set of possible fitness statistics according to the following three choices: 1) the encoding of relative abundance (e.g., transforming by a log or logit function), 2) the time scale over which to measure the change in relative abundance, and 3) the choice of a reference subpopulation for calculating relative fitness in bulk competition experiments, such as those using DNA-barcoded mutants. We show that these different choices can lead to significantly different interpretations of mutant fitness, affecting the magnitude of fitness effects, the apparent presence of epistasis, and even the fitness ranking across mutants. This can confound predictions for evolutionary dynamics and gene functions. Altogether our results demonstrate the importance of consistent fitness definitions for reproducible results across experiments.

Introduction

The concept of fitness is supposed to describe the fate of mutations as they arise in a population [1] and defines other core evolutionary concepts such as tradeoffs [2]. At the same time, fitness is often a confusing term in evolutionary biology [3–5]. It is possible to classify fitness measures by their role in theory [6] and their scale of measurement [7–9], but these arguments offer little guidance on how to measure fitness in practice. Empirical fitness measurements have become widespread in the microbial sciences — to test evolutionary theory [10–15], detect microbial interactions [16, 17], annotate gene functions [18, 19], and understand the spread of antibiotic resistance genes [20, 21] — but the approaches to quantifying fitness in these studies vary widely.

The classic approach to measuring the fitness effects of mutations uses pairwise competition experiments, where the mutant competes with its ancestral wild-type in co-culture [22–24]. This approach is ideal because it mimics the dynamics of spontaneous mutations, and thus is typically used in experimental evolution [10, 12, 25–27]. For example, the Long-Term Evolution Experiment (LTEE) has evolved Escherichia coli over tens of thousands of generations and measured the fitness of each evolved lineage in competition with the ancestor [11, 26, 28]. However, pairwise competition experiments require a way to distinguish mutants from the wild-type in cocultures (e.g., fluorescent labeling or selection markers), which is not always feasible to engineer. A second approach is thus to measure properties of genotypes in monocultures, like the growth rate, and estimate relative fitness between genotypes from these measurements. Monoculture growth curves are straightforward to measure, even for large numbers of mutants [29–31], but empirical tests have shown that the monoculture growth rate is insufficient to predict the relative fitness between genotypes [32–34]. Summary statistics of whole growth curves, like the area under the curve (AUC), perform better [35–37]. However, this approach breaks down for strains that engage in cross-feeding or toxin production, and attempts to improve monoculture-based fitness estimates require additional experiments [34, 37] and do not go beyond qualitative fitness rankings [37].

A third and more recent approach is to measure relative fitness in bulk competition experiments using high-throughput sequencing, often of DNA barcodes [13, 38]. For example, transposon-insertion mutagenesis generates libraries of gene knockouts that can be tracked as they grow in a single batch culture [39]. Estimating the fitness of each mutant allows us to identify genes that are particularly important in the growth environment [18, 19, 40], in the given genetic background [14, 15, 41, 42], and in the context of an ecological community [16, 17, 42, 43]. Bulk fitness estimation also plays a major role in identifying beneficial lineages in experimental evolution [13, 44] (not unlike the tracking of viral pathogens in public health [45–47]) and especially testing their fitness in new environments [48–52]. Many of these studies make different choices about the design of experiments and the calculation of fitness, without explaining why a certain fitness measure was used or how to compare between data sets.

The inconsistent choices for quantifying microbial fitness make it is difficult to compare fitness across experiments, especially because different groups tend to use different definitions. For example, the LTEE reports relative fitness per-generation [11, 26], whereas other evolution experiments exclusively report relative fitness percycle of batch growth [12, 53]. New techniques, like barcode sequencing, often spark their own fitness metrics [18, 40, 54–56]. A different choice of fitness metric might lead to a different ranking of mutants in a genetic screen (and the importance we assign to those genes) or change statistical properties of the distribution of mutant fitness effects, which is essential to predict the probability of mutant fixation [57], the speed of adaptation [58], and the effect size in statistical tests of gene function [18, 19]. This raises three key questions: 1) How do these fitness statistics differ? 2) Can we say which choices are optimal, in the light of first principles or some practical considerations? 3) Does this tell us how we should design these experiments? Previous work has addressed aspects of these questions [6–9, 59] but the arguments are scattered across disciplines and based on generic models of population growth [6, 7, 9, 60], rather than incorporating the specific dynamics of microbes in laboratory experiments. Here we address these questions from a unified framework, using realistic microbial population dynamics with empirical traits to systematically test different fitness metrics.

Results

Predictive power of fitness depends on the encoding of relative abundance

We distinguish between three related but distinct notions of fitness [6], all based on predicting dynamics of a population [61–64]. The first type of fitness is absolute fitness, which is a property of a single genotype by itself and serves to predict the change in the genotype’s absolute abundance N (t) over a future time window Δt (Figure S1A). This is important for questions about extinction and evolutionary rescue [65]. The second type of fitness is relative fitness, which is a property of two genotypes as it describes how the relative abundance x(t) of one genotype changes compared to the other over a time Δt (Figure S1B). This is important to determine the fixation probability of new mutations [57, 66, 67]. In general these dynamics are stochastic [66, 68, 69], but throughout this paper we focus on their average behavior across replicate cultures (as sketched in Figure S1C).

A practical challenge of working with relative fitness is that it must be measured between all pairs of genotypes in coculture competitions. Therefore it is common to infer relative fitness of two genotypes based on some individual properties of the genotypes [70–73]. We denote this third notion of fitness as the fitness potential. Like absolute fitness, fitness potential is a property of an individual genotype, but unlike the absolute fitness, it has no meaning by itself. In analogy with the potential energy of physical systems, fitness potential is intended only to be compared between two genotypes (e.g., by taking the difference or ratio) to estimate the relative fitness between those genotypes [36, 37]. The collection of fitness potential values across a large set of genotypes forms a fitness landscape [74, 75]. We note that fitness as defined here gives information about short-time dynamics but not necessarily the long-term outcome (compare Smith et al. [6]). For example, this excludes the ratio of growth rates [7, 73] or the resource concentration R∗ in chemostat equilibrium [76] since these quantities cannot tell you how fast the absolute or relative abundance is changing. We also note that in all these definitions, the fitness of a genotype is specific to one environment, including all abiotic and biotic factors. How fitness in one environment is related to fitness in other environments is important [14, 19, 42, 43, 49, 50, 73, 77, 78] but beyond the scope of this article, as we focus on specific laboratory measurements of fitness rather than abstract concepts.

In this article we will focus on relative fitness as this is the most common object of high-throughput laboratory experiments in microbes [18, 26, 40, 55, 56] and also has the most straightforward connection to key evolutionary outcomes such as fixation [57, 66, 67]. We define a genotype’s relative fitness as any number that is sufficient to predict the genotype’s relative abundance x(t) over a short time period (Figure S1). The simplest approach is to take a linear expansion of x(t) and use its slope at time t to predict the change over a time Δt (e.g., Figure 1A, top panel):

(1) x(t+Δt)=x(t)+dxdt|tΔt The slope (2) slinear=dxdt|t qualifies as a fitness statistic but often will incorrectly estimate the actual change in relative abundance because x(t) is nonlinear (Figure 1A, top panel).

Overview of the choice of encoding and the choice of time scale for quantifying relative fitness.

(A) Example trajectory of relative abundance x (top panel) for a mutant invading and eventually replacing a wild-type population. The same trajectory is plotted under the encoding log x (middle panel) and the encoding logit x = log(x/(1 − x)) (bottom panel). (B) Flowchart to predict the future relative abundance of a mutant given a relative fitness value sm = dm/dt for some encoding m. The current relative abundance xt is transformed into the new variable mt = m(xt), then projected into the future through a linear extrapolation using sm (upper horizontal arrow) and finally converted back into a frequency xt+Δt using the decoding function m−1. (C) Four scenarios for a mutant with the same relative fitness per-cycle but with different underlying population dynamics. For each scenario, we show an example trajectory of absolute abundance (stacked) for the wild-type (dark grey) and mutant population (light grey). Each scenario is mapped as a single-dot onto the fold-change diagram (center plot) and colored areas indicated positive (green area) and negative relative fitness per-cycle (blue area; compare Eq. (9)). Fitness per-cycle has isoclines that are parallel to the identity in the LFC diagram. (D) Scenario for rank discrepancy between relative fitness per-cycle scyclelogit and relative fitness per-generation sgenlogit. For a given competition (red dot), rank discrepancy occurs in a bow tie-shaped area of the fold-change space (shaded red). Any competition in the right half of this area (blue dot) will have higher mutant fitness scyclelogit but lower mutant fitness sgenlogit (right inset). As small plots on the left, we show possible population dynamics that generate this fold-change variation. Here the growth rates of the mutant and wild-type are identical in the two scenarios, but the growth period is different (e.g., from additional resources).

Alternatively, we can encode the relative abundance x in a new variable m(x) (any smooth, strictly-increasing function) whose dynamics may be easier to linearly approximate (Figure 1B; compare Denniston and Crow [79]): (3) m(x(t+Δt))=m(x(t))+smΔt, where (4) sm=dmdt|t is the relative fitness under the encoding m. In Sec. S1 we show examples of the encoded relative fitness for an explicit model of population dynamics, and in Sec. S2 we show how this concept generalizes to absolute fitness as well.

The ideal encoding transforms the relative abundance x(t) into a linear trajectory that can be fully predicted from the slope sm. While the log-encoding m(x) = log x is common in many TnSeq studies [17–19, 43, 77, 78] (and is implicit in plots of relative abundance on a logarithmic scale [44]), the log-encoded fitness slog tends to overestimate the change at high relative abundance (Figure 1A, middle panel). Instead, the best encoding in most cases is the logit function (Figure 1A, bottom panel): (5) logitx=log⁡(x1−x). This is because the relative abundance x(t) typically follows logistic dynamics (Sec. S1), at least as a first-order approximation (Sec. S3, Figure S2), and the logit function is linearly related to the inverse function of logistic x(t). The logit-encoded relative fitness of a mutant is equivalent to the standard definition of its selection coefficient in population genetics [80]. For this reason we focus on the logit-encoded relative fitness for the remainder of this work. Note that the logit encoding and logistic dynamics apply to relative abundance and are not to be confused with the logistic growth model used with microbial growth curves, where it describes the absolute abundance [81].

Estimating relative fitness requires a choice of time scale

Since the relative abundance of a genotype can only be measured at discrete times, we must estimate the instantaneous relative fitness of Eq. (4) using a finite difference between consecutive time points (Figure S1C, Sec. S4): (6) sm≈m(x(t))−m(x(t−Δt)Δt. For the logit-encoding this becomes (7) slogit=1Δt[log⁡x(t)1−x(t)−log⁡x(t−Δt)1−x(t−Δt)]. In the case of a single mutant competing against a wild-type strain, we can rewrite Eq. (7) in terms of the log fold-changes of the mutant and wild-type strains: (8) slogit=LFCmut−LFCwtΔt, where the log fold-change of each strain is LFC = log N (t)/N (t − Δt) and N is the biomass of the strain. The LFC is sometimes also referred to as a Malthusian parameter [9, 26, 65].

Therefore a key element of estimating relative fitness from discrete time data is the choice of time interval Δt. There are two common choices in empirical measurements of microbial fitness. The first is to use a time scale extrinsic to the population. In particular, microbial populations in laboratory experiments [26, 82] but also in some natural environments [83–85] grow in discrete growth cycles dictated by external pulses of nutrients (e.g., batch culture). In this case, it is convenient to quantify relative fitness per-cycle, by choosing time units in Eq. (8) such that Δt = 1: (9) scyclelogit=LFCmut−LFCwt. The second common choice for Δt is a time scale intrinsic to the population, typically its generation time. The relative fitness per-cycle may depend on the number of generations in the growth cycle, and it can be valuable to normalize for this dependence, especially when comparing across environments [26]. However, defining the number of generations for a population with multiple genotypes growing at different rates is ambiguous. In the case of a single mutant competing with a wild-type, it is common to consider only the number of generations experienced by the wild-type strain, which is estimated by its log fold-change LFCwt. It is convenient to express the generations in base e to match the natural logarithm in the logit-encoding, but this could equivalently be done by converting all logarithms in Eq. (8) to base 2. Thus the relative fitness of a mutant per-generation is defined by choosing Δt = LFCwt in Eq. (8): (10) sgenlogit=LFCmutLFCwt−1. This quantity is equivalent to the fitness statistic W=1+sgenlogit used in the LTEE [26] and other studies [86–90]. Some authors assume a fixed number of generations per growth cycle [12, 15], but since this may not be true, we use the term “per-generation” strictly for fitness statistics where the wild-type generations are explicitly measured.

Relative fitness per-cycle and per-generation can rank mutants differently when wild-type growth depends on mutant

Although relative fitness of a mutant per growth-cycle (Eq. (9)) and per-generation (Eq. (10)) differ only by choice of time scale, they are not just equivalent quantities in different units. First, while fitness per-cycle is well-defined regardless of whether the population undergoes net growth or death (Figure 1C), fitness per-generation is not defined when the wild-type undergoes no change or net death (denominator of Eq. (10) is not positive) [5, 91]. This can occur for microbes under high drug concentrations, as well as for microbial populations under harsh environmental conditions as found in sediments or wastewater [92, 93].

However, even when the wild-type LFC is positive, the relative fitness per-cycle and per-generation are not necessarily proportional to each other across a set of mutants competed against the same wild-type. The reason is that while the growth cycle time scale is extrinsic to a population and therefore the same across all pairs of genotypes, the generation time scale depends on the growth of the wild-type, which may be affected by the specific mutant it is competing against. This discrepancy can be quantitative as well as qualitative, producing different rankings of mutants. For example, Figure 1D shows two mutant genotypes (red and blue) that have opposite rankings under these fitness statistics: the blue mutant has higher relative fitness per-cycle scyclelogit, but the red mutant has higher relative fitness per-generation sgenlogit. The disagreement in ranking requires positive covariation in the LFCs (red bow tie-shaped area in Figure 1D), such that the mutant with higher LFC also induces a higher LFC in the wildtype (see illustration in Figure and exact conditions in Sec. S5.

Basic features of microbial population dynamics are sufficient to cause rank discrepancy between fitness statistics

Since the discrepancy between fitness per-cycle scyclelogit and per-generation sgenlogit occurs when the wild-type LFC varies across pairwise competitions with different mutants, we need to determine what biological mechanisms cause this. One way to explore this is to use a minimal model of microbial population dynamics parameterized by empirical data on a set of mutants. While a minimal model does not capture all possible mechanisms, it provides a null expectation for how common these discrepancies should be in real populations.

We use a model that approximates microbial population dynamics as three growth phases — a lag phase, an exponential growth phase, and a stationary phase (Figure 2A) — where strains compete for a single limiting resource with no other interactions [94, 95], since resource competition is likely to be the main interaction between a wild-type and a mutant with only small genetic differences (e.g., a point mutation or gene knockout). This model captures the minimal dynamics of microbial growth under laboratory conditions, and therefore serves as a baseline for possible outcomes. We parameterize the model using growth curves of single-gene knockout mutants in Saccharomyces cerevisiae [29, 96] (Figures and S5, Sec. S6). We simulate a pairwise competition for each knockout against the wild-type (Figure 2A) and calculate relative fitness per-cycle scyclelogit and per-generation sgenlogit for each knockout (Methods). While the relative fitness per-cycle and per-generation in the model are highly correlated across mutants overall (Figure S6), the model predicts major differences in individual mutants when we rank them by fitness (Figure 2B). Indeed, when we plot all simulated mutant competitions on the LFC diagram, we find that there are many pairs of points that have positive covariation in the mutant and wild-type LFC and thus give rise to rank differences (compare Figures 2C and 1D). Since these rankings often serve to select important mutants from a functional genomics screen or an evolution experiment, discrepancies between fitness definitions may lead analyses down different paths.

Comparison of mutant fitness rankings with different time scales using empirical trait variation.

(A) Overview of pairwise coculture simulations. For each mutant strain (grey), we simulate a competition growth cycle against a reference wild-type strain (blue) using the estimated traits and laboratory parameters for the initial condition (N0 = 0.05 OD, R0 = 111 mM glucose, x = 0.5; Methods and see Figure for the underlying trait distribution). (B) Rank discrepancy between relative fitness per-cycle scyclelogit and per-generation sgenlogit. Higher rank is defined as higher fitness, and the rank difference is defined as the rank in sgenlogit minus the rank in scyclelogit. The blue halo marks the mutant with the greatest rank difference. (C) Covariation between wild-type and mutant fold-change across all simulated competitions for all mutant strains (grey dots). The blue halo marks the same mutant as in panel B, with its red bow tie-shaped area defining the space of LFCs that would have a rank discrepancy with it (compare Figure 1D), which includes many other mutant competitions.

This model therefore shows that the basic features of microbial population dynamics and resource competition with realistic parameters are sufficient to cause substantial disagreement between fitness per-cycle and per-generation, underscoring the importance of this analysis choice. Furthermore, the model reveals a basic mechanism that causes this discrepancy: mutants must have diminished biomass yield plus variation in one additional trait (lag time or growth rate), and a high mutant relative abundance (e.g., 50%) at the start of the competition (Figure S7). These conditions are necessary because they allow mutants both to have variation in fitness and to alter the wild-type LFC in pairwise competition. Even though the wild-type growth rate is unaffected by mutants, the wild-type LFC will change in the presence of mutants that have lower yield because the population on average consumes more resources per cell division (Sec. S7). By this mechanism, it is possible to concoct a set of growth traits where relative fitness per-generation and per-cycle deliver completely opposite rankings (Figure S8).

Fitness statistics can disagree on epistasis and long-term fitness trends

Since the relative fitness of single mutants differs depending on whether fitness is quantified per-cycle or pergeneration, how does that choice impact the apparent epistatic interactions between multiple mutations and the long-term evolutionary trend in fitness? Using the same minimal model of population dynamics as in Figure 2A, we simulate a set of six mutants, each one having an increase or decrease in each of the three quantitative traits in the model (lag time, growth rate, and yield), and calculate the relative fitness per-cycle and per-generation for all pairs of those mutations (assuming the mutation effects on the traits themselves are additive; Methods). We quantify epistasis as the fitness difference (in 1:1 competition with the wild-type; Methods) between the double mutant and the sum of the single mutants (Figure 3A) [97].

Discrepancy between fitness per-generation and per-cycle in epistasis detection and long-term trend.

(A) Schematic detection of epistasis for two single mutants (light and dark circle). Epistasis occurs if the corresponding double mutant (two-component circles ) deviates from the additive fitness (parallelogram shape in solid lines; Methods). (B) Correlation between epistasis in relative fitness per-cycle scycle and per-generation sgen. Each dot corresponds to a hypothetical double mutant that combines pairs of mutations systematically perturbing each of the three growth traits (Figure 2A) in our minimal model of population dynamics (Methods). (C) Heatmap for epistasis in the per-cycle fitness scycle for all double mutants in our simulated data set, same data as in panel B. (D) Heatmap for epistasis in the per-generation fitness sgen, same data as in panel B. All epistasis plots are based on 1:1 competition growth cycles with the wild-type (Figures S9–S14). See Figure for epistasis patterns at low mutant relative abundance. (E) Hypothetical scenario for trait evolution in an evolution experiment. An evolving population decreases lag time to zero (orange line), increases growth rate until saturation (blue line), and gradually decreases biomass yield (green line). This trend is similar to initial observations from the LTEE [98]. (F) Corresponding long-term trend in relative fitness based on the trait evolution in panel E. We estimate relative fitness per-cycle scyclelogit (grey line) and per-generation sgenlogit (red line) every 250 generations. Dotted vertical lines mark the end of trait evolution in lag time and growth rate in panel E. For the actual fitness trend in the LTEE, see Figure S16.

Figure 3B shows that almost all the double mutants differ in their apparent epistasis per-cycle vs. per-generation. Epistasis per-cycle occurs when one of the mutations affects growth rate (Figure 3C), while epistasis per-generation occurs when one of the mutations affects lag time (Figure 3D). This is because the effect of a change in growth rate accrues per-generation (and hence is non-epistatic when measured over that time scale; Sec. S8), while the effect of a change in lag time accrues once per-cycle (and hence is non-epistatic in the per-cycle statistic). See Figures S9–S14 for growth curves in all cases. As an example, the double mutant in Figure S12, row D has two lag time-reducing mutations that are perfectly additive in the per-cycle statistic. However, the lag time mutations also reduce the saturation time of the growth cycle such that the fitness of the double mutant is spread over a fewer number of wild-type generations LFCwt, creating super-additive epistasis in the per-generation statistic. Note that this depends on measuring fitness for a mutant at relatively high abundance in the pairwise competition with the wild-type; when the mutant is rare, the wild-type fold-change LFCwt is effectively constant across mutants and epistasis measured per-cycle is proportional to its value measured per-generation (Figure S15). Differences in apparent epistasis can also lead to differences in apparent long-term trends of fitness per-cycle and per-generation in an evolution experiment. For example, if growth traits of a population in the minimal model evolve according to Figure 3E, then fitness per-cycle will saturate at an early time while fitness per-generation will continue to increase (Figure 3F). This is because the long-term decrease in yield reduces the wild-type LFC in 1:1 competitions between an evolved mutant and the wild-type, which causes fitness per-generation to increase (Eq. (10)).

Different fitness time scales agree on LTEE trend but not on mutant ranking

As an empirical test for discrepancies between relative fitness per-generation and per-cycle, we test how the choice of fitness time scale affects conclusions from the LTEE (Sec. S9). We reanalyze competition data from the LTEE [11] using the relative fitness per-generation sgenlogit (equivalent to the original definition used in those studies, W=1+sgenlog⁡it) and relative fitness per-cycle sgenlogit. Note these competitions are between the ancestral wild-type and each of the 12 evolved populations, rather than single mutants, across time points. We confirm that the long-term fitness trend in the LTEE is the same for scyclelogit (Figure S16). When we look more closely at the fitness ranking between lineages, we find that the two statistics disagree on the fitness ranking at almost every time point (Figure 4A) and the fittest population overall (Figure 4B; see Figure S17A for rankings). This discrepancy arises from positive covariation across pairs of LFC values in the LTEE competition data set (Figure S17B).

Rank discrepancies between fitness statistics in LTEE competition data.

(A) Rank difference between relative fitness per-cycle and per-generation as a function of time in the LTEE. For each of the 12 lines in the LTEE, we compute the relative fitness per-cycle scyclelogit and per-generation sgenlogit as a function of time, by averaging the data from the LTEE competition data set [11] across replicate measurements (Sec. S9). For some lines, the time-series is truncated due to measurement difficulties [11]. Based on the quantitative fitness values scyclelogit and sgenlogit, we compute fitness rankings between the replicate lines at any given time point (higher rank means higher fitness ; compare Figure 2B). The rank difference is defined as the rank in sgenlogit minus the rank in scyclelogit, and we combine the rank difference between all pairs of evolving lines to compute the maximum rank difference at each time-point. Insets: At three chosen time-points (t = 4000, t = 15000, t = 30000 generations), we show the correlation between relative fitness per-cycle scyclelogit and relative fitness per-generation sgenlogit (each dot corresponds to one of the 12 lines in the LTEE). We quantify the correlation between scyclelogit and sgenlogit with the Spearman rank correlation λ (panel title). Colors indicate the time-point and correspond to the color bar in panel B. (B) Covariation between relative fitness per-cycle scyclelogit and per-generation sgenlogit for evolved populations of the LTEE. Each line of the LTEE contributes roughly two competitions per time-point due to replicate measurements (Sec. S9) [11]. The color of the points indicates evolutionary time point; see panel (C) for the color bar. We highlight the evolved population with the greatest rank difference (blue halo; same as in Figure S17). (C) Same as panel B but comparing relative fitness under the logit-encoding scyclelogit and under the log-encoding scyclelog.

While we have focused on the logit encoding so far since logistic dynamics are usually the best null model of mutant frequency dynamics (Figure 1A), the log encoding is common in many TnSeq studies [17–19, 43, 77, 78]. The LTEE competition data is a valuable opportunity to empirically test the consequences of this choice as well. Figure 4C shows that the relative fitness per-cycle under the log-encoding scyclelog differs from the logit-based fitness scyclelogit for evolved LTEE populations, leading to ranking discrepancies especially for the fittest populations (Figure S18A). The discrepancy in fitness between encodings has a different mechanism compared to the discrepancy from fitness time scale: it arises from variation in the initial relative abundance in mutant-to-wild-type competitions in the LTEE data set. Although these pairwise competitions are intended to start with equal abundances of the evolved population and of the wild-type (1:1), Figure S18B shows that there is a systematic trend toward lower initial frequencies of the evolved populations over time. This trend is a side effect of the inoculation method in LTEE competition experiments, where a fixed volume of the evolved population is transferred to the competition growth cycle but the cell density for the evolved population in pre-culture decreases [98, 99].

Fitness in bulk competitions is distorted by higher-order effects depending on reference subpopulation

So far we have focused on relative fitness in pairwise competition experiments, but this is often not practical for large numbers of mutants due to the required strain labeling (e.g., with fluorescent proteins or selection markers). One alternative is to measure a fitness potential (as defined at the beginning of the Results section), such as growth rate or area under the growth curve (AUC) [32, 34, 36, 37], from monoculture experiments, under the assumption that true relative fitness is proportional to the difference in this quantity between genotypes. Using our minimal model of population dynamics parameterized by single-gene knockouts in yeast (Figure 2A), we find that common fitness potentials perform poorly overall (Figure S19; see refs. [32, 34, 36, 37] for empirical tests), with some success for the AUC but strongly depending on the choice of time window (Figure S20, Sec. S10).

Another approach to measuring relative fitness for a large number of mutants is to use bulk competition experiments, where many mutant genotypes compete simultaneously in a single culture and each genotype is tracked through high-throughput sequencing [39, 44]. This raises the question of how well relative fitness of a mutant in bulk competition corresponds to its relative fitness in a pairwise competition with the wild-type, since the growth in bulk might be influenced by the presence of other mutant genotypes, a phenomenon known as a higher-order interaction [78, 95, 100].

An important choice in bulk competition experiments is the relative abundance of the library of all mutant genotypes compared to the wild-type. The invasion of a spontaneous mutation into an existing population is best captured by pairwise competitions with low initial relative abundance of the mutant (Case I, Figure 5A), and one way to recreate this scenario in a bulk competition is to use a low relative abundance of the mutant library overall (Case II, Figure 5A) [40, 48–52]. A practical problem with Case II is that individual genotypes in the library will have low absolute abundances, which leads to stochasticity in the population dynamics and the sequencing process [13, 101, 102]. Therefore, it is common to compete the mutant library by itself (Case III, Figure 5A) [13, 15, 17–19, 77].

The choice of library abundance and reference group in bulk competition experiments.

(A) Overview of a pairwise competition experiment with one mutant (top row) and scenarios for bulk competition experiments with many different mutants (middle and bottom rows). The two bulk competition scenarios differ in their initial fraction of the mutant library (colored ovals) in the inoculum (open box). For each scenario, we show a schematic growth cycle (log absolute abundance) in the inset on the right. (B) Schematic relative abundance trajectories for a mutant compared to two alternative subpopulations. We distinguish between the total relative abundance xi with respect to the population as a whole (height of green band in the top box) and the pairwise relative abundance xi,wt with respect to the wild-type (height of green band in the bottom box; Eq. (20)). We indicate the sign of total relative fitness (Eq. (24)) and pairwise relative fitness (Eq. (25)) on the right. (C) The absolute error between bulk and pairwise competition experiments. The total relative fitness (dark grey dots; Eq. (24)) and the pairwise relative fitness (orange dots; Eq. (25)) for mutants in the minimal population dynamics model (Figure 2A, Methods) parameterized by the yeast knockout data (Figure S4) in bulk competition growth cycle with low mutant library abundance (panel A, case II; Methods). The absolute error is defined as the bulk fitness statistic minus the relative fitness in pairwise competition (Eq. (S92) in Sec. S15). In the inset, the absolute error for pairwise relative fitness (Eq. (25)) for a bulk competition growth cycle with 99.9% library abundance (light grey dots; case III). This still includes a wild-type reference at small percentage to be able to compute a pairwise relative fitness. The x-axis and the orange dots in the inset are identical to the main plot. (D) The relative error in bulk competition experiments as a function of mutant library abundance in the inoculum. Each line corresponds to a knockout mutant in our data set, and represents the relative error between the pairwise relative fitness in bulk competition and the relative fitness in pairwise competition (Eq. (S102) in Sec. S16). The black vertical lines show the recommended mutant library abundance for our data set based on Eq. (11) (xlib ≈ 24.6%) and based on the more conservative Eq. (S118) (xlib ≈ 0.02%, Sec. S16). Note these bounds are calculated a priori and are not the result of a fitting procedure.

A second important choice is whether to quantify fitness of a mutant relative to the whole population (“total relative fitness”) [18, 19, 77] or relative to another specific genotype, like the wild-type (“pairwise relative fitness”; see Figure 5B) [40, 48–52]. In the case of a mutant library growing by itself (Case III), it is still possible to estimate a pairwise fitness by using a few known neutral mutants as the reference population [15, 17] and calculating fitness relative to this group (Methods; Secs. and S12).

To test the consequences of these choices, we use the minimal population dynamics model (Figure 2A) parameterized by yeast single-gene knockouts and compare the estimates of total and pairwise relative fitness in bulk competitions to fitness measured in pairwise competition, which we use as the ground truth (Methods). Figure 5C shows that in bulk competitions, the error in total relative fitness (compared to fitness in pairwise competitions) is systematically offset from the error in pairwise relative fitness. For example, a mutant that grows identically to the wild-type (neutral phenotype) does not change in abundance relative to the wild-type, but it may have a net increase in its total relative abundance due to the poor growth of other mutants (compare top and bottom panel in Figure 5B). This offset affects the total relative fitness of all mutants in a uniform way (Figure 5C), such that total and pairwise relative fitness completely agree in the ranking of genotypes (Figure S21). Note that this off-set between total and relative fitness is general and does not depend on our minimal population dynamics model (Sec. S13).

While the pairwise relative fitness is the preferred method for estimating fitness in bulk competitions, the presence of higher-order interactions means it still deviates from the fitness in pairwise competitions. Figure 5C shows the absolute error for pairwise fitness has a positive trend with fitness and increases with the relative abundance of the mutant (Case III, inset in Figure 5C). For the specific population dynamics in our simulation, we decompose the higher-order interaction into two terms (Secs. and S15): a fitness-independent term and a fitness-dependent term that acts as an amplifier for fitness values (Figure S22) and drives the positive error trend in Figure 5C. Intuitively, the population in bulk competition consumes resources more slowly and gives more time for growth rate differences to accrue. The amplification of mutant pairwise fitness in bulk is important because it widens the apparent distribution of fitness effects (leading to, for example, higher estimates for the speed of adaptation) than if fitness was measured in pairwise competition.

In order to control the error from higher-order inter-actions, we derive an explicit error bound. Figure 5D shows the relative error from higher-order interactions for each mutant across a range of mutant library abundances. Using the specific population dynamics of our model, we derive the following estimate: Assuming that the mutants in the library only have variation in growth rate, the mutant library abundance should satisfy (11) xlib<ϵ⋅gwt|glib−gwt|, where ε is the desired threshold on the relative error (Figure S23), gwt is the wild-type growth rate, and glib is the growth rate of the library as a whole (Sec. S16). In the case of our single-gene knockout library, Eq. (11) predicts a maximum library abundance of 24.6% based on the wild-type and library growth rate (gwt = 0.406, glib = 0.389) for a relative error ε = 1%. Figure 5D shows that this maximum library abundance keeps the relative error below 1% for high-fitness mutants (bright yellow), because they are dominated by growth rate effects, but fails for mutants close to neutrality because they have a tradeoff between growth rate and lag time, which the estimate in Eq. (11) neglects. It is possible to derive a more precise bound that keeps the relative error below 1% for all mutant genotypes (in this case xlib = 0.01%, see vertical line in Figure 5D), but this requires prior knowledge of the trait covariation in the mutant library (Sec. S16).

Previous benchmark experiments show deviation of bulk fitness from pairwise competition

Our models predict that higher-order effects amplify pairwise fitness measured in bulk compared to the fitness in a pairwise competition experiment. To test this prediction, we reanalyze two published benchmark experiments of bulk fitness measurements. The first data set by van Opijnen et al. [103] measured pairwise (1:1) and bulk competitions (xlib ≈ 99.9%) of 46 single-gene knockouts in Streptococcus pneumoniae using logit-encoded fitness per-generation and neutral strains as the reference group. Figure 6A shows that the fitness in bulk deviates from the diagonal, and the absolute error has a positive correlation with the underlying true fitness from pairwise competition (Figure S24A; Pearson r = 0.40,p = 0.006), similar to the pattern observed in our simulations (compare Figure S24A with Figure 5C case III, and Figure S24B with Figure S23C case III).

Empirical benchmarks of bulk fitness against pairwise competition experiments.

(A) Fitness measurements as reported by Opijnen et al. [103] for knockouts of S. pneumoniae. Pairwise fitness was measured in 1:1 competitions of knockouts strains with the wild-type over a single growth cycle using the per-generation statistic W=1+sgenlogit [26]. Bulk fitness was measured from a single growth cycle using 100% initial abundance of the mutant library and a set of neutral strains as the reference to calculate a per-generation fitness, replacing LFCwt with the library LFC in Eq. (S31) (Sec. S5). Since we were unable to obtain the original data, we extracted these fitness data from Figure 3c in the original publication and subtracted 1 to plot as sgenlogit. See Figure S24 for the absolute and relative error of bulk fitness based on these data. (B) Fitness measurements as reported by Levy et al. [13] for evolved genotypes of of S. cerevisiae after 80 generations of the evolution experiment. Pairwise fitness was measured in 1:1 competitions of knockouts trains with a fluorescently-labeled wild-type over seven growth cycle using relative fitness per-cycle under the log-encoding scyclelog, divided by a fixed number generations inferred from the dilution factor. Bulk fitness was measured from relative abundance data throughout the evolution experiment and using a multi-step algorithm, Bayesian inference, and a set of neutral strains as the reference, leading to a pairwise fitness estimate equivalent to scyclelogit (compare SI Eq. (50) in Levy et al. [13]). We extracted the fitness data from the table presented in SI Sec. 8.1 of the original publication. See Figure for the absolute and relative error of bulk fitness based on these data.

The second data set by Levy et al. [13] measured pairwise competitions (1:1) and relative abundances for 33 lab-evolved genotypes of S. cerevisiae with beneficial fitness. The bulk fitness is inferred under the log-encoding and per-cycle, with neutral strains as the reference group. Figure 6B shows that the fitness of strains in the bulk evolution environment is typically higher than in pair-wise competition, but the absolute error has no positive correlation (Figure S25A; Pearson r = −0.26,p = 0.140) and thus does not support the pattern we see our simulations (compare Figure S25A with Figure Figure case III, and S25B with Figure S23C case III).

Discussion

Significance of fitness quantification choices for microbial ecology and evolution

In this work, we have introduced a conceptual framework to derive common statistics of relative fitness from three essential choices:

The choice of the state variable for relative abundance (encoding m).

The choice of the time scale for the change in relative abundance (Δt).

In bulk competition experiments, the choice of subpopulation that acts as the reference for relative abundance of a chosen mutant genotype (Figure 5B).

The combination of these choices leads to a range of fitness statistics, including those commonly used in population genetics [80], experimental evolution [11, 12, 26], and gene knockout screens [17–19, 42, 43, 77, 78, 103].

However, as we compare these statistics in simulated competition experiments, we find that the choice of time scale can lead to different mutant rankings (Figures 2B and 4A,B), confound apparent epistasis between mutations (Figure 3B), and alter the apparent long-term trend in fitness (Figure 3F), while the choice of the reference subpopulation can lead to a systematic shift in inferred fitness values (Figure 5C). The quantitative differences between relative fitness statistics affect our ability to make evolutionary predictions. In particular, the fixation probability of a spontaneous mutation depends on the magnitude of relative fitness compared to other time scales like mutation and drift [66, 67], and in microbial populations with clonal interference, it depends on the entire distribution of fitness effects [58, 104]. The choice of statistic may also affect the outcome of a significance test in the context of gene essentiality tests (e.g., using log or logit with the test statistic in Wetmore et al. [18]). Since these measurements often serve as a first screen to narrow down the investigation to the top set of genes [17, 40], a difference in mutant ranking means the investigation might miss relevant genes because of the choice of fitness statistic.

Best practices for quantifying mutant fitness in high-throughput experiments

Based on these insights, we recommend the following choices for quantifying relative fitness of a mutant:

Use the logit encoding of relative abundance, because under the null model of logistic dynamics this linearizes the trajectory of relative abundance. Only use a different encoding if alternate, non-logistic dynamics are known.

For batch growth or any experiment with non-steady state growth, use a fixed extrinsic time scale (e.g., a single growth cycle), rather than an intrinsic time scale (e.g., the number of generations), which introduces additional variation between competitions. Only quantify fitness per generation for experiments that are known to have steady-state growth (e.g., in a chemostat).

In bulk competition experiments, always use the pairwise fitness relative to the wild-type (by including a barcoded wild-type or by grouping neutral mutants into a virtual wild-type) because this more closely matches the relative fitness in pairwise competition.

To control the error from higher-order interactions, we recommend minimizing relative abundance of the mutant library as practically feasible. For a given error tolerance, Eq. (11) gives an estimate for the maximum library abundance.

Our recommendations agree with previous criticisms of relative fitness per-generation [5, 8, 9, 91] and total relative fitness [59, 105] but differ from the standard practice in high-throughput genetic screens [18, 19, 77] and many evolution experiments [33, 88], including the LTEE [11, 26]. The choice of the reference group is a well-recognized issue in high-throughput evolutionary studies, in which one first estimates a total relative fitness and then subtracts a correction based on the fitness of the wild-type [12, 50, 52] or a mean population fitness [13, 41, 42, 51, 105]. In contrast, we recommend to choose the reference group at the level of relative abundance (see Methods for a step-by-step demonstration). While researchers must ultimately choose how to quantify fitness based on the specific goals and conditions of their experiment, we believe that a more consistent adherence to the recommendations above would be valuable to the field as a whole so that fitness data from one study can be quantitatively compared to fitness from another.

How general is the disagreement between fitness statistics in pairwise competition?

The discrepancy in mutant ranking for relative fitness per-cycle scyclelogit and per-generation sgenlogit requires positive covariation between the wild-type and mutant log fold-changes (LFCs) across a set of pairwise competition experiments. This condition is not dependent on any particular mechanism or model, although our minimal model of population dynamics (Figure 2A) demonstrates one way this covariation can occur: when mutants have diminished yields and are present at sufficiently-high abundances to affect wild-type LFCs (Figure S7).

Our model focused on the effect of resource competition on fitness discrepancies since that interaction is almost always present between microbial genotypes, providing a baseline expectation. Moreover, resource competition is likely to be the only interaction in high-throughput fitness measurements of mutants with only small genetic differences (e.g., point mutations or gene knockouts) relative to a wild-type. We expect that in cases with additional interactions between mutants and the wild-type, there will be further variation in wild-type LFCs across competitions and thus even more discrepancies betwen fitness statistics. For example, mutants that stimulate wild-type growth via the secretion of nutrients should create positive LFC covariation across mutants and thus be ranked differently per-cycle and per-generation. Indeed, our reanalysis of LTEE competition data shows that fitness per-cycle scyclelogit and per-generation sgenlogit (equivalent to the statistic W=1+sgenlogit used in the LTEE) disagree on the fitness ranking at almost every time point (Figure 4A) and in the first 50,000 generations overall (Figure 4B).

How general is the deviation of bulk fitness statistics from pairwise competition experiments?

The offset in total relative fitness (between grey and orange points in Figure 5C), equivalent to the mean library fitness (Eq. (S88) in Sec. S13), is independent from the population dynamics (Figure 5B, Sec. S13). In contrast, the error between pairwise fitness in bulk competition and in pairwise competition (orange points in Figure 5C) is specific to the population dynamics studied here, although it points to more general mechanisms. Specifically, we expect pairwise fitness in bulk competition to deviate from pairwise competition as soon as the mutant library is able to influence the chemical environment of the competition, for example, when a large fraction of mutants fails to complete respiration of the main carbon source and instead turns to fermentation [106–108], with the release of pathway intermediates like acetate that indirectly lower the pH in the growth medium [109, 110].

The choice of fitness statistic in the light of other sources of discrepancy

Besides the conceptual choices of fitness quantification discussed in this article, experimental limitations can create discrepancies between replicate fitness measurements of the same mutant [36, 50, 111]. For example, relative abundance measurements entail sampling uncertainties (when we sample liquid for colony counting or DNA sequencing) [44, 59, 60], copy number variation [18, 44], as well as PCR jackpots and sequencing read errors [13, 101, 102]. Furthermore, there will inevitably be some variation in initial condition between replicates and fluctuations during the fitness assay [36, 102, 111].

Methods

Inferring growth traits for the single-gene knockout collection in yeast

We use a previously published data set [29] for the single-gene knockout collection in Saccharomyces cerevisiae [96], where the authors track growth of each genotype for 47 hours in monoculture, using microwell plates with defined growth medium. We download this growth curve data set from the PROPHECY database [112], choosing specifically the data set measured in Synthetic Defined medium. We last accessed the PROPHECY website (http://prophecy.lundberg.gu.se/) on March 30, 2020, but as of November 15, 2025 the website is no longer accessible. We have included a reformatted version of the data with our code repository (https://github.com/justuswfink/24FitnessQuantification). From the raw time series of optical density, the authors subtracted a background correction based on blank wells, applied another correction for nonlinearity at high optical densities, and smoothed the growth curve to remove electrical noise [29].

We start with the curves in the published data set (9951 curves) and apply further trimming and smoothing steps (Sec. S6). From this data we calculate the time series of instantaneous per-capita growth rate N−1dN/dt (where N is the optical density) and identify time windows where the rate is approximately constant, which we interpret as distinct growth phases (Sec. S6). We only include curves that have a single phase of constant exponential growth followed by a stationary phase of approximately zero growth (9424 curves).

We quantify the biomass yield, maximum growth rate, and the lag time from each remaining curve as follows. First, we estimate the initial abundance Ninitial (average optical density over first three time points) and the final abundance Nfinal (average optical density over stationary phase) from the growth curve. Then we calculate the biomass yield as Y = (Nfinal − Ninitial)/R(0), where R(0) = 111 mM (20 g/L) is the initial concentration of glucose (assuming glucose is the single limiting resource). To estimate the maximum growth rate, we average the instantaneous growth rate over the exponential phase. Finally, we estimate the lag time from the intersection of the log initial abundance (log N0) with the slope of the maximum growth rate during the exponential growth phase [81]. We excluded curves with negative initial OD, curves with negative inferred lag times, and curves with a low quality of fit between the measured time series and a simulated curve based on the inferred trait values (R2 < 0.95; see below for the model). Our final database includes trait estimates for 9195 curves, which represents 92.4% of the original data set [29].

Each single-gene deletion strain was measured in two technical replicate growth curves, using a second plate with identical layout in the same plate reader. Some genotypes have only one estimate of the growth traits because the other replicate did not pass our filters (273 genotypes), but for most genotypes we retain two replicate estimates in our final data set (4163 genotypes); a few genotypes even have three (2 genotypes) or four replicate estimates (54 genotypes) because these geno-types were included multiple times by the original authors [29]. From these replicates we finally calculate the average yield, growth rate, and lag time for each geno-type. The data set also contained many replicate growth curves of the wild-type strain, 374 of which passed our filters. Since wild-type traits inferred from these replicates had large variation (potentially due to measuring the wild-type across many different plates and days), we define the wild-type trait as the median, not the mean, of these replicates. Although the wild-type measurements included in this data set have large trait variation (orange dots in Figure S4B,C), the fact that the variation across knockouts is significantly greater than for the wild-type replicates in terms of growth rate (Levene’s test p = 1.3 × 10−9) and lag time (Levene’s test p = 0.025), and that the replicate measurements of gene knockouts are correlated (Figure S5), suggests that these trait measurements do capture true genetic variation. In contrast, the variation in biomass yield across knockouts is not significantly different from the variation across wild-type replicates (Levene’s test p = 0.44), suggesting that the variation in biomass yield is driven by non-genetic factors like the abundance or batch of resources used for each growth measurement.

Simulating population dynamics in competition experiments

Throughout this work, we model the competition between a set of genotypes during a single batch culture growth cycle using the following model of population dynamics [94, 95]: (12a) 1NjdNjdt=gj(t,R),Nj(0)=xj(0)N0 (12b) dRdt=−∑j1YjdNjdt,R(0)=R0 where for each genotype j, Nj is the absolute abundance, xj(0) the initial relative abundance, gj(t, R) the growth rate, and Yj the biomass yield, where the genotypes include a wild-type as well as one or more mutants. The initial absolute abundance of all genotypes together is N0, and the concentration of the single limiting resource is R with initial value R0. Note that Eq. (12b) assumes that cells consume resources only for biomass growth and not for maintenance of existing biomass.

To capture the sigmoidal shape of typical growth curves with a lag phase, exponential growth phase, and saturation phase (e.g., Figure S4A), we model the dependence of growth rate on time and resource concentration as (13) gj(t,R)=gjΘ(t−λj)Θ(R), where Θ is the step function: (14) Θ(z)={0z≤0,1z>0. In this model, a genotype has an initial lag phase of time λj where no growth occurs, followed by a phase of constant exponential growth at rate gj, and ending when the resource concentration R reaches zero. The time to resource depletion, which we also call the saturation time, is defined by the implicit equation R(tsat) = 0. The simple form of the growth response (Eq. (13)) means that the absolute abundance of genotype j at saturation is (15) Nj(tsat)=Nj(t0)egj(tsat−λj)Θ(tsat−λj) and its log fold-change is (16) LFCj=gj(tsat−λj)Θ(tsat−λj). To calculate relative fitness per-cycle (Eq. (9) and per-generation (Eq. (10)), we numerically determine the saturation time tsat by integrating the biomass and resource dynamics (Eq. (12)) as described in previous work [94, 95]. We note that it also possible to derive an approximate expression for the saturation time tsat which we use for our analytical calculations (Sec. S7).

Testing epistasis between mutations in population dynamics model

Consider a wild-type strain with growth rate gwt = 1, lag time λwt = 2, and biomass yield Ywt = 1. We define six basic mutation that represent the six directions of the trait space in our model: a mutation that increases growth rate (gmut = 1.2; symbol +g), a mutation that decreases growth rate (gmut = 0.8; symbol −g), a mutation that increases lag time (λmut = 3; symbol +λ), a mutation that decreases lag time (λmut = 1; symbol −λ), a mutation that increases yield (Ymut = 4; symbol +Y ), and a mutation that decreases yield (Ymut = 0.25; symbol −Y ). We combine these mutations into double mutants as follows: mutations are additive at the level of traits (e.g., the double mutant (+g, −g) is neutral) and mutations in different traits are independent. For a given pair of mutations, epistasis is defined as [97] (17) E=s(A,B)−s(A)−s(B), where s(A) is the relative fitness of the single mutant with mutation A, s(B) is the relative fitness of the single mutant with mutation B, and s(A, B) is the relative fitness of the double mutant with both A and B. We estimate the fitness of each mutant by simulating one growth cycle of competition with the wild-type (at initially equal abundance) using an initial biomass of N0 = 0.01 and initial resource concentration R0 = 0.5. These conditions are chosen such that mutations that increase biomass yield reach a maximum density of up to N (tsat) = 2, on the same scale as the optical density measured for the empirical knockouts of S. cerevisiaea (Figure S4C).

Quantifying total and pairwise relative fitness in bulk competition experiments

We define total and pairwise relative fitness,two alternative measures of fitness in bulk competition experiments, as follows. Consider a population with multiple genotypes such that each genotype i has absolute abundance Ni(t) at time t. The total relative abundance of the genotype is (18) xi(t)=Ni(t)∑jNj(t). The “total” here refers to the fact that this is the abundance of genotype i relative to all other genotypes in the population. Similarly, we can think of the definition of relative fitness discussed in the main text (Eq. (4)) as the total relative fitness of this genotype under an encoding (19) σim=ddtm(xi(t)). However, sometimes we want to track the dynamics of a genotype relative to another specific genotype; for example, to follow the fate of a mutant against the wild-type in a bulk competition experiment. For a pair of genotypes i ≠ j, we thus define the pairwise relative abundance (20) xij(t)=xi(t)xi(t)+xj(t), which is the relative abundance of genotype i in a subpopulation of only genotypes i and j. To predict the change in the pairwise relative abundance, we can use the slope (21) sijm=ddtm(xij(t)), which defines the pairwise relative fitness of genotype i with respect to genotype j (see Sec. for a generalize definition that covers the case i = j). In the special case of a population with only two strains, the pair-wise relative fitness and total relative fitness are identical (σ2m=s21m) but they may differ with more than two geno-types (Sec. S13).

It is also possible to define the total and pairwise relative fitness statistics as finite differences over a time interval, rather than as instantaneous derivatives. For example, we can define them over a growth cycle starting at t = 0 and ending at t = tsat: (22) σi,cyclem=m(xi(tsat))−m(xi(0)), (23) sij,cyclem=m(xij(tsat))−m(xij(0)). For these fitness statistics in the bulk competitions, we use the logit encoding (24) σi,cyclelogit=log⁡xi(tsat)1−xi(tsat)−log⁡xi(0)1−xi(0), (25) sij,cyclelogit=log⁡xi(tsat)xi(0)−log⁡xj(tsat)xj(0), where we have rewritten sij,cyclelogit in the form that it is presented in bulk competition experiments [15, 51, 52]. The logit encoding has mathematical advantages for coarse-graining the relative fitness of genotype groups (Sec. S11) but using the log encoding is another common choice in the literature [17, 41, 50]. However, the relative abundances of the individual mutants in our simulated bulk competition experiments are low enough that these two encodings are approximately equivalent (logit x ≈ log x for x ≈ 1).

Unlike the comparison between per-cycle and per-generation relative fitness where we focused on rank differences (Figure 2B), here we can evaluate absolute differences in fitness estimates because the total and pairwise fitness in bulk fitness are measured in the same units as the relative fitness in pairwise competition. For all mutant genotypes, we calculate the absolute error between these bulk fitness estimates (Eqs. (24) and (25)) and the relative fitness in pairwise competitions (Eq. (9)), which we take as the ground truth (Figure 5C). Note that the pairwise competition could depend on the initial relative abundance of the mutant; we have chosen a very low relative abundance (10−6) that mimics a mutant arising de novo and where this dependence is very weak.

Clearly, minimizing the abundance of the mutant library minimizes the strength of higher-order interactions (also reducing ranking disagreement, Figure S21B,C). As low mutant abundances create practical problems for experiments, it is valuable to identify a maximum abundance for the mutant library that keeps the error from bulk fitness estimates below a desired threshold (Eq. (11)). It is convenient to express this threshold in relative rather than absolute terms (compare Figure S23 to Figure 5C).

A guide to calculate pairwise relative fitness under the logit encoding from bulk competition data

Here ri(t) represents the best estimate of read numbers for lineage i after correcting for experimental artifacts, like jackpot events in the sequencing (e.g., see discussion in Levy et al. [13]).

Calculate the total relative abundance xi of each lineage i by dividing with the total number of reads: (26) xi(t)=ri(t)∑i=1nri(t), where the sum is over all n lineages in the experiment.

Define a group of lineages i = 1, . .., k as the reference (e.g., a group of neutral mutants) and sum their total relative abundance (27) xref(t)=∑i=1kxi(t).

Calculate the pairwise relative abundance xi,ref of each lineage by (28) xi,ref(t)=xi(t)xi(t)+xref(t).

Calculate the pairwise fitness per-cycle for each strain i under the logit encoding (29) si,reflogit=1Δt[log⁡xi,ref(t2)1−xi,ref(t2)−log⁡xi,ref(t1)1−xi,ref(t1)], where Δt is the number of growth cycles between time points t1 and t2.

eLife Assessment

This manuscript applies a theoretical analysis to two published datasets on yeast and bacterial evolution to compare different ways of quantifying fitness. It makes an important advance by clarifying how discrepancies can arise by using different approaches and provides recommendations for best practices. Overall, this is an impressive and highly beneficial study that is based on convincing evidence and has the potential of setting standards in this rapidly growing field.

Reviewer #1 (Public review):

The authors point out that the fitness estimates obtained from different experimental assays (monoculture, pairwise competition or bulk competition) are not generally equivalent, not even with regard to the fitness ranking of different genotypes. Using a computational model based on experimentally measured growth phenotypes for knockout strains in yeast, as well as data from Lenski's Long Term Evolution Experiment (LTEE), they derive a set of best practice rules aimed at extracting the optimal amount of information from such experiments.

The study is very complete on a technical level, and the conceptual weaknesses raised in the first round of reviews have been fully addressed in the revision.

Reviewer #2 (Public review):

Summary:

The manuscript "Quantifying microbial fitness in high-throughput experiments" provides a comprehensive analysis of the various approaches to quantifying fitness in microbial evolution, focusing on three primary factors: encoding of relative abundance, time scale of measurement, and the choice of reference subpopulation. The authors systematically explore how these choices impact fitness statistics and provide recommendations aimed at standardizing practices in the field. This manuscript aims to highlight the impact of differing fitness definitions and the methodologies utilized for analysis and how that can significantly alter interpretations of mutant fitness, affecting evolutionary predictions and the overall understanding of genetic interactions in the experiments.

Strengths:

The choices for quantifying fitness in evolution experiments are critical and highly relevant given the increasing prevalence of high-throughput experiments in evolutionary biology. The authors methodically categorize fitness statistics and their implications, providing clarity on a complex subject. This structured approach aids in understanding the nuances of fitness measurement. The manuscript effectively highlights how different choices in fitness measurement can influence fitness rankings and the understanding of epistasis, which is important for modeling evolutionary dynamics.

Comments on revisions:

The authors have comprehensively addressed all previous comments and suggestions. In particular, the addition of the new methods section: 'A guide to calculate pairwise relative fitness under the logit encoding from bulk competition data' - significantly improves the clarity of the implementation and helps in the overall interpretation of the framework.

Reviewer #3 (Public review):

Summary:

The authors present analyses of different fitness measures derived from empirical data from yeast knock-out mutants and the long-term evolution experiment (LTEE) with Escherichia coli to explore discrepancies and identify preferred methods to estimate relative fitness in high-throughput experiments. Their work has three components. They first discuss the different "encodings" of relative abundance data and conclude that logit-transformations are preferred, because they transform nonlinear abundance trajectories into linear trajectories with greater predictive power. Next, they compare per-generation with per-growth cycle relative fitness estimates inferred from simulations of pairwise competitions based on published growth traits for the yeast strains and on published pairwise competition measurements for the LTEE data. Both data sets show quantitative and qualitative (i.e. rank order) discrepancies of estimates across different time scales, which are highlighted by considering possible underlying causes (i.e. trade-offs between growth traits) and consequences (i.e. epistasis among mutations affecting different growth traits). Finally, the authors compare simulated pairwise and bulk (i.e. where many mutants compete during a growth cycle in a single environment) competition assays based on the yeast knock-out mutants and demonstrate an optimal ratio of collective mutants to wild-type strains that minimizes both sampling error and overestimation of fitness estimates when compared with pairwise competitions.

Strengths:

The study deals with a highly relevant topic. Fitness is central to general evolutionary theory, but also poorly defined and implies different traits for different organisms and conditions. For microbes, which are often used in evolution experiments, high-throughput experiments may yield different measures to quantify abundance over time, from individual growth traits to bulk competition experiments. Hence, it is relevant to consider discrepancies among those measures and identify preferred measures with respect to predicting population dynamic and evolutionary processes. The present study contributes to this aim by (i) making readers aware of differences among commonly used fitness estimates, (ii) showing that simulated (yeast) and calculated (E. coli) competitive fitness may differ across time scales, and (iii) showing that bulk competitions may yield relative fitness estimates that are systematically higher than pairwise competitions. The study is rather thorough on the theory side, with extensive derivations and analyses of various fitness measures using their resource competition model in the Supplementary Information. The study ends with a few practical recommendations for preferred methods to infer relative fitness estimates, that may be useful for experimentalists and stimulate further investigations.

Weaknesses:

The study has a few limitations. Perhaps the most apparent limitation is the lack of a clear answer to the question which fitness measure is best "in the light of first principles". The authors show clear discrepancies between fitness estimates across different time scales or using different reference genotypes in bulk competition and provide useful recommendations based on practical considerations (e.g. using pairwise competitions as "golden standard"), but it remains unclear whether these measures provide the greatest value for the questions researchers may want to answer with them (e.g. predict shifts in genotype frequencies). -- The authors have convinced me in their response that their recommendations were fundamentally related to the resource competition model, and the changes in introduction and discussion help to appreciate the choice of fitness measure in relation to the research question.

A second limitation is that the authors analyse fitness differences arising solely from resource competition, whereas microbes often interact via other mechanisms, e.g. the production of anticompetitor toxins, cross-feeding of metabolites or lack of growth to enhance their persistence in stress conditions. Without simulations of these processes, understanding discrepancies among fitness measures is necessarily limited. In addition, the analysis of trade-offs between growth traits causing these discrepancies during resource competition seems confounded by biases in measurement error or parameter estimation, at least for growth rate and lag time (Fig. 2B), where the replicate estimates for the wildtype show a similar negative correlation. -- The motivation to use a resource competition model for fitness inference is generally well motivated now. I accept their argument that resource competitive differences are most important for microbial strains with small genetic differences (e.g. from mutant libraries or from the same evolution experiment). However, it is relevant to note that this ignores situations that are rather common, where the wild-type strain produces an anticompetitor toxin or causes growth inhibition through metabolite products that lower the pH (and derived strains will likely contain resistant mutations).

Third, the study does not validate relative fitness predictions from growth traits (as is done for the yeast mutants) with measured relative fitness estimates using competition assays, while such data are available, e.g. for the LTEE. This would strengthen their inferences about preferred fitness measures. -- In their response, the authors explain that their aim was different, i.e. the provide "proof of principle" that the choices of fitness measure can produce discrepancies even when they follow the same growth model.

Fourth, the analysis of epistasis between mutations affecting different growth traits (shown in Fig. 3) based on the LTEE data could be better introduced and analysed more comprehensively. Now, the examples given in panels C-F seem rather idiosyncratic and readers may wonder how general these consequences of using fitness estimates based on different time scales are. -- The authors have made extensive improvements to address how different growth parameters, especially lag and growth rate, differently affect apparent epistasis based on measures at different time scale (per generation vs per cycle). These provide a more comprehensive analysis of down-stream consequences for epistasis detection.

Finally, the study is generally less accessible to experimentalists due to the extensive and principled treatment of specific population dynamic models and fitness inferences. This may distract from the overarching aim to identify fitness measures that are most accurate and useful for predictions of population dynamic and evolutionary processes. In this light, the motivation for the initial discussion of the importance of how to best encode relative abundance (Fig. 1) is unclear. Also, the conclusion, that logit encoding is preferred, because it linearizes logistic growth dynamics and "improves the quality of predictions", is not further motivated. Experimentalists using non-linear models to infer fitness from growth curves or competition assays may miss the relevance of this discussion. -- Thanks for this explanation (indeed, I confused "logistic dynamics" with "logistic growth model"); the additional explanations and text reductions have improved accessibility for experimentalists.

Comments on revisions:

I appreciate the thorough and effective response to all recommendations and have no further comments.

Author response:

The following is the authors’ response to the previous reviews

We thank both editors and the three reviewers for their positive feedback on the revised manuscript. Following suggestions from Reviewer #1, we cite additional literature to clarify the use of ’fitness potential’ and we have revised the caption of Figure 3 to better explain how we generate the panel of mutants. We have also added a sentence to emphasize that the selection coefficient should match the time-scale of bottleneck effects in the evolution environment.

Version 3

Abstract

Few concepts are as central to evolution as is fitness, and yet the quantification of fitness is often ambiguous. In particular, high-throughput experiments to measure mutant fitness in microbes are increasingly common but vary widely in their definitions of fitness, which makes their results difficult to compare. What are the consequences of these different fitness statistics, and is there a best way to quantify fitness in a given context? Here we systematize the set of possible fitness statistics according to the following three choices: 1) the encoding of relative abundance (e.g., transforming by a log or logit function), 2) the time scale over which to measure the change in relative abundance, and 3) the choice of a reference subpopulation for calculating relative fitness in bulk competition experiments, such as those using DNA-barcoded mutants. We show that these different choices can lead to significantly different interpretations of mutant fitness, affecting the magnitude of fitness effects, the apparent presence of epistasis, and even the fitness ranking across mutants. This can confound predictions for evolutionary dynamics and gene functions. Altogether our results demonstrate the importance of consistent fitness definitions for reproducible results across experiments.

Introduction

The concept of fitness is supposed to describe the fate of mutations as they arise in a population [1] and defines other core evolutionary concepts such as tradeoffs [2]. At the same time, fitness is often a confusing term in evolutionary biology [3–5]. It is possible to classify fitness measures by their role in theory [6] and their scale of measurement [7–9], but these arguments offer little guidance on how to measure fitness in practice. Empirical fitness measurements have become widespread in the microbial sciences — to test evolutionary theory [10–15], detect microbial interactions [16, 17], annotate gene functions [18, 19], and understand the spread of antibiotic resistance genes [20, 21] — but the approaches to quantifying fitness in these studies vary widely.

The classic approach to measuring the fitness effects of mutations uses pairwise competition experiments, where the mutant competes with its ancestral wild-type in co-culture [22–24]. This approach is ideal because it mimics the dynamics of spontaneous mutations, and thus is typically used in experimental evolution [10, 12, 25–27]. For example, the Long-Term Evolution Experiment (LTEE) has evolved Escherichia coli over tens of thousands of generations and measured the fitness of each evolved lineage in competition with the ancestor [11, 26, 28]. However, pairwise competition experiments require a way to distinguish mutants from the wild-type in cocultures (e.g., fluorescent labeling or selection markers), which is not always feasible to engineer. A second approach is thus to measure properties of genotypes in monocultures, like the growth rate, and estimate relative fitness between genotypes from these measurements. Monoculture growth curves are straightforward to measure, even for large numbers of mutants [29–31], but empirical tests have shown that the monoculture growth rate is insufficient to predict the relative fitness between genotypes [32–34]. Summary statistics of whole growth curves, like the area under the curve (AUC), perform better [35–37]. However, this approach breaks down for strains that engage in cross-feeding or toxin production, and attempts to improve monoculture-based fitness estimates require additional experiments [34, 37] and do not go beyond qualitative fitness rankings [37].

A third and more recent approach is to measure relative fitness in bulk competition experiments using high-throughput sequencing, often of DNA barcodes [13, 38]. For example, transposon-insertion mutagenesis generates libraries of gene knockouts that can be tracked as they grow in a single batch culture [39]. Estimating the fitness of each mutant allows us to identify genes that are particularly important in the growth environment [18, 19, 40], in the given genetic background [14, 15, 41, 42], and in the context of an ecological community [16, 17, 42, 43]. Bulk fitness estimation also plays a major role in identifying beneficial lineages in experimental evolution [13, 44] (not unlike the tracking of viral pathogens in public health [45–47]) and especially testing their fitness in new environments [48–52]. Many of these studies make different choices about the design of experiments and the calculation of fitness, without explaining why a certain fitness measure was used or how to compare between data sets.

The inconsistent choices for quantifying microbial fitness make it is difficult to compare fitness across experiments, especially because different groups tend to use different definitions. For example, the LTEE reports relative fitness per-generation [11, 26], whereas other evolution experiments exclusively report relative fitness percycle of batch growth [12, 53]. New techniques, like barcode sequencing, often spark their own fitness metrics [18, 40, 54–56]. A different choice of fitness metric might lead to a different ranking of mutants in a genetic screen (and the importance we assign to those genes) or change statistical properties of the distribution of mutant fitness effects, which is essential to predict the probability of mutant fixation [57], the speed of adaptation [58], and the effect size in statistical tests of gene function [18, 19]. This raises three key questions: 1) How do these fitness statistics differ? 2) Can we say which choices are optimal, in the light of first principles or some practical considerations? 3) Does this tell us how we should design these experiments? Previous work has addressed aspects of these questions [6–9, 59] but the arguments are scattered across disciplines and based on generic models of population growth [6, 7, 9, 60], rather than incorporating the specific dynamics of microbes in laboratory experiments. Here we address these questions from a unified framework, using realistic microbial population dynamics with empirical traits to systematically test different fitness metrics.

Results

Predictive power of fitness depends on the encoding of relative abundance

We distinguish between three related but distinct notions of fitness [6], all based on predicting dynamics of a population [61–64]. The first type of fitness is absolute fitness, which is a property of a single genotype by itself and serves to predict the change in the genotype’s absolute abundance N (t) over a future time window Δt (Fig. S1A). This is important for questions about extinction and evolutionary rescue [65]. The second type of fitness is relative fitness, which is a property of two geno-types as it describes how the relative abundance x(t) of one genotype changes compared to the other over a time Δt (Fig. S1B). This is important to determine the fixation probability of new mutations [57, 66, 67]. In general these dynamics are stochastic [66, 68, 69], but throughout this paper we focus on their average behavior across replicate cultures (as sketched in Fig. S1C).

A practical challenge of working with relative fitness is that it must be measured between all pairs of genotypes in coculture competitions. Therefore it is common to infer relative fitness of two genotypes based on some individual properties of the genotypes [70–73]. We denote this third notion of fitness as the fitness potential. Like absolute fitness, fitness potential is a property of an individual genotype, but unlike the absolute fitness, it has no meaning by itself. In analogy with the potential energy of physical systems, fitness potential is intended only to be compared between two genotypes (e.g., by taking the difference or ratio) to estimate the relative fitness between those genotypes [36, 37]. The collection of fitness potential values across a large set of genotypes forms a fitness landscape [74, 75]. Note that the term “fitness potential” has been used with a related but somewhat different meaning in the population genetics literature [76–78].

We emphasize that fitness as defined here gives information about short-time dynamics but not necessarily the long-term outcome (compare Smith et al. [6]). For example, this excludes the ratio of growth rates [7, 73] or the resource concentration R ∗ in chemostat equilibrium [79] since these quantities cannot tell you how fast the absolute or relative abundance is changing. We also note that in all these definitions, the fitness of a genotype is specific to one environment, including all abiotic and biotic factors. How fitness in one environment is related to fitness in other environments is important [14, 19, 42, 43, 49, 50, 73, 80, 81] but beyond the scope of this article, as we focus on specific laboratory measurements of fitness rather than abstract concepts.

In this article we will focus on relative fitness as this is the most common object of high-throughput laboratory experiments in microbes [18, 26, 40, 55, 56] and also has the most straightforward connection to key evolutionary outcomes such as fixation [57, 66, 67]. We define a genotype’s relative fitness as any number that is sufficient to predict the genotype’s relative abundance x(t) over a short time period (Fig. S1). The simplest approach is to take a linear expansion of x(t) and use its slope at time t to predict the change over a time Δt (e.g., Fig. 1A, top panel):

Overview of the choice of encoding and the choice of time scale for quantifying relative fitness.

(A) Example trajectory of relative abundance x (top panel) for a mutant invading and eventually replacing a wild-type population. The same trajectory is plotted under the encoding log x (middle panel) and the encoding logit x = log(x/(1 − x)) (bottom panel). (B) Flowchart to predict the future relative abundance of a mutant given a relative fitness value sm = dm/dt for some encoding m. The current relative abundance xt is transformed into the new variable mt = m(xt), then projected into the future through a linear extrapolation using sm (upper horizontal arrow) and finally converted back into a frequency xt+!t using the decoding function m−1. (C) Four scenarios for a mutant with the same relative fitness per-cycle but with different underlying population dynamics. For each scenario, we show an example trajectory of absolute abundance (stacked) for the wild-type (dark grey) and mutant population (light grey). Each scenario is mapped as a single-dot onto the fold-change diagram (center plot) and colored areas indicated positive (green area) and negative relative fitness per-cycle (blue area; compare Eq. (9)). Fitness per-cycle has isoclines that are parallel to the identity in the LFC diagram. (D) Scenario for rank discrepancy between relative fitness per-cycle scyclelogit and relative fitness per-generation sgenlogit. For a given competition (red dot), rank discrepancy occurs in a bow tie-shaped area of the fold-change space (shaded red). Any competition in the right half of this area (blue dot) will have higher mutant fitness scyclelogit but lower mutant fitness sgenlogit (right inset). As small plots on the left, we show possible population dynamics that generate this fold-change variation. Here the growth rates of the mutant and wild-type are identical in the two scenarios, but the growth period is different (e.g., from additional resources).

(1) x(t+Δt)=x(t)+dxdt|tΔt. The slope (2) slinear=dxdt|t qualifies as a fitness statistic but often will incorrectly estimate the actual change in relative abundance because x(t) is nonlinear (Fig. 1A, top panel).

Alternatively, we can encode the relative abundance x in a new variable m(x) (any smooth, strictly-increasing function) whose dynamics may be easier to linearly approximate (Fig. 1B; compare Denniston and Crow [82]): (3) m(x(t+Δt))=m(x(t))+smΔt, where (4) sm=dmdt|t is the relative fitness under the encoding m. In Sec. S1 we show examples of the encoded relative fitness for an explicit model of population dyanmics, and in Sec. S2 we show how this concept generalizes to absolute fitness as well.

The ideal encoding transforms the relative abundance x(t) into a linear trajectory that can be fully predicted from the slope sm. While the log-encoding m(x) = log x is common in many TnSeq studies [17–19, 43, 80, 81] (and is implicit in plots of relative abundance on a logarithmic scale [44]), the log-encoded fitness slog tends to overestimate the change at high relative abundance (Fig. 1A, middle panel). Instead, the best encoding in most cases is the logit function (Fig. 1A, bottom panel): (5) logitx=log⁡(x1−x) This is because the relative abundance x(t) typically follows logistic dynamics (Sec. S1), at least as a first-order approximation (Sec. S3, Fig. S2), and the logit function is linearly related to the inverse function of logistic x(t). The logit-encoded relative fitness of a mutant is equivalent to the standard definition of its selection coefficient in population genetics [83]. For this reason we focus on the logit-encoded relative fitness for the remainder of this work. Note that the logit encoding and logistic dynamics apply to relative abundance and are not to be confused with the logistic growth model used with microbial growth curves, where it describes the absolute abundance [84].

Estimating relative fitness requires a choice of time scale

Since the relative abundance of a genotype can only be measured at discrete times, we must estimate the instantaneous relative fitness of Eq. (4) using a finite difference between consecutive time points (Fig. S1C, Sec. S4): (6) sm≈m(x(t))−m(x(t−Δt)Δt. For the logit-encoding this becomes (7) slogit=1Δt[log⁡x(t)1−x(t)−log⁡x(t−Δt)1−x(t−Δt)]. In the case of a single mutant competing against a wild-type strain, we can rewrite Eq. (7) in terms of the log fold-changes of the mutant and wild-type strains: (8) slogit=LFCmut−LFCwtΔt, where the log fold-change of each strain is LFC = log N (t)/N (t − Δt) and N is the biomass of the strain. The LFC is sometimes also referred to as a Malthusian parameter [9, 26, 65].

Therefore a key element of estimating relative fitness from discrete time data is the choice of time interval Δt. There are two common choices in empirical measurements of microbial fitness. The first is to use a time scale extrinsic to the population. In particular, microbial populations in laboratory experiments [26, 85] but also in some natural environments [86−88] grow in discrete growth cycles dictated by external pulses of nutrients (e.g., batch culture). In this case, it is convenient to quantify relative fitness per-cycle, by choosing time units in Eq. (8) such that Δt = 1: (9) scyclelogit=LFCmut−LFCwt. The second common choice for Δt is a time scale intrinsic to the population, typically its generation time. The relative fitness per-cycle may depend on the number of generations in the growth cycle, and it can be valuable to normalize for this dependence, especially when comparing across environments [26]. However, defining the number of generations for a population with multiple genotypes growing at different rates is ambiguous. In the case of a single mutant competing with a wild-type, it is common to consider only the number of generations experienced by the wild-type strain, which is estimated by its log fold-change LFCwt. It is convenient to express the generations in base e to match the natural logarithm in the logit-encoding, but this could equivalently be done by converting all logarithms in Eq. (8) to base 2. Thus the relative fitness of a mutant per-generation is defined by choosing Δt = LFCwt in Eq. (8): (10) sgenlogit=LFCmutLFCwt−1. This quantity is equivalent to the fitness statistic W=1+sgenlogit used in the LTEE [26] and other studies [89–93]. Some authors assume a fixed number of generations per growth cycle [12, 15], but since this may not be true, we use the term “per-generation” strictly for fitness statistics where the wild-type generations are explicitly measured.

Relative fitness per-cycle and per-generation can rank mutants differently when wild-type growth depends on mutant

Although relative fitness of a mutant per growth-cycle (Eq. (9)) and per-generation (Eq. (10)) differ only by choice of time scale, they are not just equivalent quantities in different units. First, while fitness per-cycle is well-defined regardless of whether the population undergoes net growth or death (Fig. 1C), fitness per-generation is not defined when the wild-type undergoes no change or net death (denominator of Eq. (10) is not positive) [5, 94]. This can occur for microbes under high drug concentrations, as well as for microbial populations under harsh environmental conditions as found in sediments or wastewater [95, 96].

However, even when the wild-type LFC is positive, the relative fitness per-cycle and per-generation are not necessarily proportional to each other across a set of mutants competed against the same wild-type. The reason is that while the growth cycle time scale is extrinsic to a population and therefore the same across all pairs of genotypes, the generation time scale depends on the growth of the wild-type, which may be affected by the specific mutant it is competing against. This discrepancy can be quantitative as well as qualitative, producing different rankings of mutants. For example, Fig. 1D shows two mutant genotypes (red and blue) that have opposite rankings under these fitness statistics: the blue mutant has higher relative fitness per-cycle scyclelogit, but the red mutant has higher relative fitness per-generation sgenlogit. The disagreement in ranking requires positive covariation in the LFCs (red bow tie-shaped area in Fig. 1D), such that the mutant with higher LFC also induces a higher LFC in the wild-type (see illustration in Fig. and exact conditions in Sec.S5)

Basic features of microbial population dynamics are sufficient to cause rank discrepancy between fitness statistics

Since the discrepancy between fitness per-cycle scyclelogit and per-generation sgenlogit occurs when the wild-type LFC varies across pairwise competitions with different mutants, we need to determine what biological mechanisms cause this. One way to explore this is to use a minimal model of microbial population dynamics parameterized by empirical data on a set of mutants. While a minimal model does not capture all possible mechanisms, it provides a null expectation for how common these discrepancies should be in real populations.

We use a model that approximates microbial population dynamics as three growth phases — a lag phase, an exponential growth phase, and a stationary phase (Fig. 2A) — where strains compete for a single limiting resource with no other interactions [97, 98], since resource competition is likely to be the main interaction between a wild-type and a mutant with only small genetic differences (e.g., a point mutation or gene knockout). This model captures the minimal dynamics of microbial growth under laboratory conditions, and therefore serves as a baseline for possible outcomes. We parameterize the model using growth curves of single-gene knockout mutants in Saccharomyces cerevisiae [29, 99] (Figs. and S5, Sec. S6). We simulate a pairwise competition for each knockout against the wild-type (Fig. 2A) and calculate relative fitness per-cycle scyclelogit and per-generation sgenlogit for each knockout (Methods). While the relative fitness per-cycle and per-generation in the model are highly correlated across mutants overall (Fig. S6), the model predicts major differences in individual mutants when we rank them by fitness (Fig. 2B). Indeed, when we plot all simulated mutant competitions on the LFC diagram, we find that there are many pairs of points that have positive covariation in the mutant and wild-type LFC and thus give rise to rank differences (compare Figs. 2C and 1D). Since these rankings often serve to select important mutants from a functional genomics screen or an evolution experiment, discrepancies between fitness definitions may lead analyses down different paths.

Comparison of mutant fitness rankings with different time scales using empirical trait variation.

(A) Overview of pairwise coculture simulations. For each mutant strain (grey), we simulate a competition growth cycle against a reference wild-type strain (blue) using the estimated traits and laboratory parameters for the initial condition (N0 = 0.05 OD, R0 = 111 mM glucose, x = 0.5; Methods and see Fig. for the underlying trait distribution). (B) Rank discrepancy between relative fitness per-cycle scyclelogit and per-generation sgenlogit. Higher rank is defined as higher fitness, and the rank difference is defined as the rank in sgenlogit minus the rank in scyclelogit. The blue halo marks the mutant with the greatest rank difference. (C) Covariation between wild-type and mutant fold-change across all simulated competitions for all mutant strains (grey dots). The blue halo marks the same mutant as in panel B, with its red bow tie-shaped area defining the space of LFCs that would have a rank discrepancy with it (compare Fig. 1D), which includes many other mutant competitions.

This model therefore shows that the basic features of microbial population dynamics and resource competition with realistic parameters are sufficient to cause substantial disagreement between fitness per-cycle and pergeneration, underscoring the importance of this analysis choice. Furthermore, the model reveals a basic mechanism that causes this discrepancy: mutants must have diminished biomass yield plus variation in one additional trait (lag time or growth rate), and a high mutant relative abundance (e.g., 50%) at the start of the competition (Fig. S7). These conditions are necessary because they allow mutants both to have variation in fitness and to alter the wild-type LFC in pairwise competition. Even though the wild-type growth rate is unaffected by mutants, the wild-type LFC will change in the presence of mutants that have lower yield because the population on average consumes more resources per cell division (Sec. S7). By this mechanism, it is possible to concoct a set of growth traits where relative fitness per-generation and per-cycle deliver completely opposite rankings (Fig. S8).

Fitness statistics can disagree on epistasis and long-term fitness trends

Since the relative fitness of single mutants differs depending on whether fitness is quantified percycle or per-generation, how does that choice impact the apparent epistatic interactions between multiple mutations and the long-term evolutionary trend in fitness? Using the same minimal model of population dynamics as in Fig. 2A, we simulate a set of six mutants, each one having an increase or decrease in each of the three quantitative traits in the model (lag time, growth rate, and yield), and calculate the relative fitness per-cycle and per-generation for all pairs of those mutations (assuming the mutation effects on the traits themselves are additive; Methods). We quantify epistasis as the fitness difference (in 1:1 competition with the wild-type; Methods) between the double mutant and the sum of the single mutants (Fig. 3A) [100].

Discrepancy between fitness per-generation and per-cycle in epistasis detection and long-term trend.

(A) Schematic detection of epistasis for two single mutants (light and dark circle). Epistasis occurs if the corresponding double mutant (two-component circles) deviates from the additive fitness (parallelogram shape in solid lines; Methods). (B) Correlation between epistasis in relative fitness per-cycle scycle and per-generation sgen. Each dot corresponds to a hypothetical double mutant that combines pairs of mutations systematically perturbing each of the three growth traits (Fig. 2A) in our minimal model of population dynamics (Methods). (C) Heatmap for epistasis in the per-cycle fitness scycle for all double mutants in our simulated data set, same data as in panel B. The rows and columns correspond to the six single mutations in our model of population dynamics (Methods): an increase or decrease in lag time (+ω, − ω), an increase or decrease in maximum growth rate (+g, − g), or an increase or decrease in biomass yield (+Y, −Y). (D) Heatmap for epistasis in the per-generation fitness sgen, same data as in panel B. All epistasis plots are based on 1:1 competition growth cycles with the wild-type (Figs. –). See Fig. for epistasis patterns at low mutant relative abundance. (E) Hypothetical scenario for trait evolution in an evolution experiment. An evolving population decreases lag time to zero (orange line), increases growth rate until saturation (blue line), and gradually decreases biomass yield (green line). This trend is similar to initial observations from the LTEE [101]. (F) Corresponding long-term trend in relative fitness based on the trait evolution in panel E. We estimate relative fitness per-cycle Scycle (grey line) and per-generation sgenlogit (red line) every 250 generations. Dotted vertical lines mark the end of trait evolution in lag time and growth rate in panel E. For the actual fitness trend in the LTEE, see Fig.

Figure 3B shows that almost all the double mutants differ in their apparent epistasis per-cycle vs. per-generation. Epistasis per-cycle occurs when one of the mutations affects growth rate (Fig. 3C), while epistasis pergeneration occurs when one of the mutations affects lag time (Fig. 3D). This is because the effect of a change in growth rate accrues per-generation (and hence is non-epistatic when measured over that time scale; Sec. S8), while the effect of a change in lag time accrues once per-cycle (and hence is non-epistatic in the percycle statistic). See Figs. S9–S14 for growth curves in all cases. As an example, the double mutant in Fig. S12, row D has two lag time-reducing mutations that are perfectly additive in the per-cycle statistic. However, the lag time mutations also reduce the saturation time of the growth cycle such that the fitness of the double mutant is spread over a fewer number of wild-type generations LFCwt, creating super-additive epistasis in the per-generation statistic. Note that this depends on measuring fitness for a mutant at relatively high abundance in the pairwise competition with the wild-type; when the mutant is rare, the wild-type fold-change LFCwt is effectively constant across mutants and epistasis measured per-cycle is proportional to its value measured per-generation (Fig. S15). Differences in apparent epistasis can also lead to differences in apparent long-term trends of fitness per-cycle and per-generation in an evolution experiment. For example, if growth traits of a population in the minimal model evolve according to Fig. 3E, then fitness per-cycle will saturate at an early time while fitness per-generation will continue to increase (Fig. 3F). This is because the long-term decrease in yield reduces the wild-type LFC in 1:1 competitions between an evolved mutant and the wild-type, which causes fitness per-generation to increase (Eq. (10)).

Different fitness time scales agree on LTEE trend but not on mutant ranking

As an empirical test for discrepancies between relative fitness per-generation and per-cycle, we test how the choice of fitness time scale affects conclusions from the LTEE (Sec. S9). We reanalyze competition data from the LTEE [11] using the relative fitness per-generation sgenlogit (equivalent to the original definition used in those studies, W=1+sgenlogit) and relative fitness per-cycle scyclelogit. Note these competitions are between the ancestral wild-type and each of the 12 evolved populations, rather than single mutants, across time points. We confirm that the long-term fitness trend in the LTEE is the same for scyclelogit (Fig. S16). When we look more closely at thefitness ranking between lineages, we find that the two statistics disagree on the fitness ranking at almost every time point (Fig. 4A) and the fittest population overall (Fig. 4B; see Fig. S17A for rankings). This discrepancy arises from positive covariation across pairs of LFC values in the LTEE competition data set (Fig. S17B).

Rank discrepancies between fitness statistics in LTEE competition data.

(A) Rank difference between relative fitness per-cycle and per-generation as a function of time in the LTEE. For each of the 12 lines in the LTEE, we compute the relative fitness per-cycle Scycle and per-generation sgenlogit as a function of time, by averaging the data from the LTEE competition data set [11] across replicate measurements (Sec. S9). For some lines, the time-series is truncated due to measurement difficulties [11]. Based on the quantitative fitness values Scycle and sgenlogit, we compute fitness rankings between the replicate lines at any given time point (higher rank means higher fitness; compare Fig. 2B). The rank difference is defined as the rank in sgenlogit minus the rank in scyclelogit, and we combine the rank difference between all pairs of evolving lines to compute the maximum rank difference at each time-point. Insets: At three chosen time-points (t = 4000, t = 15000, t = 30000 generations), we show the correlation between relative fitness per-cycle Scycle and relative fitness per-generation sgenlogit (each dot corresponds to one of the 12 lines in the LTEE). We quantify the correlation between scyclelogit and sgenlogit with the Spearman rank correlation ε (panel title). Colors indicate the time-point and correspond to the color bar in panel B. (B) Covariation between relative fitness per-cycle scyclelogit and per-generation sgenlogit for evolved populations of the LTEE. Each line of the LTEE contributes roughly two competitions per time-point due to replicate measurements (Sec. S9) [11]. The color of the points indicates evolutionary time point; see panel (C) for the color bar. We highlight the evolved population with the greatest rank difference (blue halo; same as in Fig. S17). (C) Same as panel B but comparing relative fitness under the logit-encoding scyclelogit and under the log-encoding scyclelog.

While we have focused on the logit encoding so far since logistic dynamics are usually the best null model of mutant frequency dynamics (Fig. 1A), the log encoding is common in many TnSeq studies [17–19, 43, 80, 81]. The LTEE competition data is a valuable opportunity to empirically test the consequences of this choice as well. Figure 4C shows that the relative fitness per-cycle under the log-encoding scyclelog differs from the logit-based fitness scyclelogit for evolved LTEE populations, leading to ranking discrepancies especially for the fittest populations (Fig. S18A). The discrepancy in fitness between encodings has a different mechanism compared to the discrepancy from fitness time scale: it arises from variation in the initial relative abundance in mutant-to-wild-type competitions in the LTEE data set. Although these pairwise competitions are intended to start with equal abundances of the evolved population and of the wild-type (1:1), Fig. S18B shows that there is a systematic trend toward lower initial frequencies of the evolved populations over time. This trend is a side effect of the inoculation method in LTEE competition experiments, where a fixed volume of the evolved population is transferred to the competition growth cycle but the cell density for the evolved population in pre-culture decreases [101, 102].

Fitness in bulk competitions is distorted by higher-order effects depending on reference subpopulation

So far we have focused on relative fitness in pairwise competition experiments, but this is often not practical for large numbers of mutants due to the required strain labeling (e.g., with fluorescent proteins or selection markers). One alternative is to measure a fitness potential (as defined at the beginning of the Results section), such as growth rate or area under the growth curve (AUC) [32, 34, 36, 37], from monoculture experiments, under the assumption that true relative fitness is proportional to the difference in this quantity between genotypes. Using our minimal model of population dynamics parameterized by single-gene knockouts in yeast (Fig. 2A), we find that common fitness potentials perform poorly overall (Fig. S19; see refs. [32, 34, 36, 37] for empirical tests), with some success for the AUC but strongly depending on the choice of time window (Fig. S20, Sec. S10).

Another approach to measuring relative fitness for a large number of mutants is to use bulk competition experiments, where many mutant genotypes compete simulta-neously in a single culture and each genotype is tracked through high-throughput sequencing [39, 44]. This raises the question of how well relative fitness of a mutant in bulk competition corresponds to its relative fitness in a pairwise competition with the wild-type, since the growth in bulk might be influenced by the presence of other mu-tant genotypes, a phenomenon known as a higher-order interaction [81, 98, 103].

An important choice in bulk competition experiments is the relative abundance of the library of all mutant genotypes compared to the wild-type. The invasion of a spontaneous mutation into an existing population is best captured by pairwise competitions with low initial relative abundance of the mutant (Case I, Fig. 5A), and one way to recreate this scenario in a bulk competition is to use a low relative abundance of the mutant library overall (Case II, Fig. 5A) [40, 48–52]. A practical problem with Case II is that individual genotypes in the library will have low absolute abundances, which leads to stochasticity in the population dynamics and the sequencing process [13, 104, 105]. Therefore, it is common to compete the mutant library by itself (Case III, Fig. 5A) [13, 15, 17–19, 80].

The choice of library abundance and reference group in bulk competition experiments.

(A) Overview of a pairwise competition experiment with one mutant (top row) and scenarios for bulk competition experiments with many different mutants (middle and bottom rows). The two bulk competition scenarios differ in their initial fraction of the mutant library (colored ovals) in the inoculum (open box). For each scenario, we show a schematic growth cycle (log absolute abundance) in the inset on the right. (B) Schematic relative abundance trajectories for a mutant compared to two alternative subpopulations. We distinguish between the total relative abundance xi with respect to the population as a whole (height of green band in the top box) and the pairwise relative abundance xi,wt with respect to the wild-type (height of green band in the bottom box; Eq. (20)). We indicate the sign of total relative fitness (Eq. (24)) and pairwise relative fitness (Eq. (25)) on the right. (C) The absolute error between bulk and pairwise competition experiments. The total relative fitness (dark grey dots; Eq. (24)) and the pairwise relative fitness (orange dots; Eq. (25)) for mutants in the minimal population dynamics model (Fig. 2A, Methods) parameterized by the yeast knockout data (Fig. S4) in bulk competition growth cycle with low mutant library abundance (panel A, case II; Methods). The absolute error is defined as the bulk fitness statistic minus the relative fitness in pairwise competition (Eq. (S92) in Sec. S15). In the inset, the absolute error for pairwise relative fitness (Eq. (25)) for a bulk competition growth cycle with 99.9% library abundance (light grey dots; case III). This still includes a wild-type reference at small percentage to be able to compute a pairwise relative fitness. The x-axis and the orange dots in the inset are identical to the main plot. (D) The relative error in bulk competition experiments as a function of mutant library abundance in the inoculum. Each line corresponds to a knockout mutant in our data set, and represents the relative error between the pairwise relative fitness in bulk competition and the relative fitness in pairwise competition (Eq. (S102) in Sec. S16). The black vertical lines show the recommended mutant library abundance for our data set based on Eq. (11) (xlib ≈ 24.6%) and based on the more conservative Eq. (S118) (xlib ≈ 0.02%, Sec. S16). Note these bounds are calculated a priori and are not the result of a fitting procedure.

A second important choice is whether to quantify fitness of a mutant relative to the whole population (“total relative fitness”) [18, 19, 80] or relative to another specific genotype, like the wild-type (“pairwise relative fitness”; see Fig. 5B) [40, 48–52]. In the case of a mutant library growing by itself (Case III), it is still possible to estimate a pairwise fitness by using a few known neutral mutants as the reference population [15, 17] and calculating fitness relative to this group (Methods; Secs. and S12).

To test the consequences of these choices, we use the minimal population dynamics model (Fig. 2A) parameterized by yeast single-gene knockouts and compare the estimates of total and pairwise relative fitness in bulk competitions to fitness measured in pairwise competition, which we use as the ground truth (Methods). Figure 5C shows that in bulk competitions, the error in total relative fitness (compared to fitness in pairwise competitions) is systematically offset from the error in pairwise relative fitness. For example, a mutant that grows identically to the wild-type (neutral phenotype) does not change in abundance relative to the wild-type, but it may have a net increase in its total relative abundance due to the poor growth of other mutants (compare top and bottom panel in Fig. 5B). This offset affects the total relative fitness of all mutants in a uniform way (Fig. 5C), such that total and pairwise relative fitness completely agree in the ranking of genotypes (Fig. S21). Note that this offset between total and relative fitness is general and does not depend on our minimal population dynamics model (Sec. S13).

While the pairwise relative fitness is the preferred method for estimating fitness in bulk competitions, the presence of higher-order interactions means it still deviates from the fitness in pairwise competitions. Figure 5C shows the absolute error for pairwise fitness has a positive trend with fitness and increases with the relative abundance of the mutant (Case III, inset in Fig. 5C). For the specific population dynamics in our simulation, we decompose the higher-order interaction into two terms (Secs. and S15): a fitness-independent term and a fitness-dependent term that acts as an amplifier for fitness values (Fig. S22) and drives the positive error trend in Fig. 5C. Intuitively, the population in bulk competition consumes resources more slowly and gives more time for growth rate differences to accrue. The amplification of mutant pairwise fitness in bulk is important be-cause it widens the apparent distribution of fitness effects (leading to, for example, higher estimates for the speed of adaptation) than if fitness was measured in pairwise competition.

In order to control the error from higher-order interactions, we derive an explicit error bound. Figure 5D shows the relative error from higher-order interactions for each mutant across a range of mutant library abundances. Using the specific population dynamics of our model, we derive the following estimate: Assuming that the mutants in the library only have variation in growth rate, the mutant library abundance should satisfy (11) xlib<ϵ⋅gwt|glib−gwt|, where ω is the desired threshold on the relative error (Fig. S23), gwt is the wild-type growth rate, and glib is the growth rate of the library as a whole (Sec. S16). In the case of our single-gene knockout library, Eq. (11) predicts a maximum library abundance of 24.6% based on the wild-type and library growth rate (gwt = 0.406, glib = 0.389) for a relative error ω = 1%. Figure 5D shows that this maximum library abundance keeps the relative error below 1% for high-fitness mutants (bright yellow), because they are dominated by growth rate effects, but fails for mutants close to neutrality because they have a tradeoff between growth rate and lag time, which the estimate in Eq. (11) neglects. It is possible to derive a more precise bound that keeps the relative error below 1% for all mutant genotypes (in this case xlib = 0.01%, see vertical line in Fig. 5D), but this requires prior knowledge of the trait covariation in the mutant library (Sec. S16).

Previous benchmark experiments show deviation of bulk fitness from pairwise competition

Our models predict that higher-order effects amplify pairwise fitness measured in bulk compared to the fitness in a pairwise competition experiment. To test this prediction, we reanalyze two published benchmark experiments of bulk fitness measurements. The first data set by van Opijnen et al. [106] measured pairwise (1:1) and bulk competitions (xlib ≈ 99.9%) of 46 single-gene knockouts in Streptococcus pneumoniae using logit-encoded fitness per-generation and neutral strains as the reference group. Figure 6A shows that the fitness in bulk deviates from the diagonal, and the absolute error has a positive correlation with the underlying true fitness from pairwise competition (Fig. S24A; Pearson r = 0.40,p = 0.006), similar to the pattern observed in our simulations (compare Fig. S24A with Fig. 5C case III, and Fig. S24B with Fig. S23C case III).

Empirical benchmarks of bulk fitness against pairwise competition experiments.

(A) Fitness measurements as reported by Opijnen et al. [106] for knockouts of S. pneumoniae. Pairwise fitness was measured in 1:1 competitions of knockouts strains with the wild-type over a single growth cycle using the per-generation statistic W=1+sgenlogit [26]. Bulk fitness was measured from a single growth cycle using 100% initial abundance of the mutant library and a set of neutral strains as the reference to calculate a per-generation fitness, replacing LFCwt with the library LFC in Eq. (S31) (Sec. S5). Since we were unable to obtain the original data, we extracted these fitness data from Fig. 3c in the original publication and subtracted 1 to plot as sgenlogit. See Fig. S24 for the absolute and relative error of bulk fitness based on these data. (B) Fitness measurements as reported by Levy et al. [13] for evolved genotypes of of S. cerevisiae after 80 generations of the evolution experiment. Pairwise fitness was measured in 1:1 competitions of knockouts trains with a fluorescently-labeled wild-type over seven growth cycle using relative fitness per-cycle under the log-encoding scyclelog, divided by a fixed number generations inferred from the dilution factor. Bulk fitness was measured from relative abundance data throughout the evolution experiment and using a multi-step algorithm, Bayesian inference, and a set of neutral strains as the reference, leading to a pairwise fitness estimate equivalent to scyclelogit (compare SI Eq. (50) in Levy et al. [13]). We extracted the fitness data from the table presented in SI Sec. 8.1 of the original publication. See Fig. for the absolute and relative error of bulk fitness based on these data.

The second data set by Levy et al. [13] measured pair-wise competitions (1:1) and relative abundances for 33 lab-evolved genotypes of S. cerevisiae with beneficial fitness. The bulk fitness is inferred under the log-encoding and per-cycle, with neutral strains as the reference group. Figure 6B shows that the fitness of strains in the bulk evolution environment is typically higher than in pair-wise competition, but the absolute error has no positive correlation (Fig. S25A; Pearson r = −0.26,p = 0.140) and thus does not support the pattern we see our simulations (compare Fig. S25A with Fig. 5 case III, and Fig. S25B with Fig. S23C case III).

Discussion

Significance of fitness quantification choices for microbial ecology and evolution

In this work, we have introduced a conceptual framework to derive common statistics of relative fitness from three essential choices:

The choice of the state variable for relative abundance (encoding m).

The choice of the time scale for the change in relative abundance (Δt).

In bulk competition experiments, the choice of sub-population that acts as the reference for relative abundance of a chosen mutant genotype (Fig. 5B).

The combination of these choices leads to a range of fitness statistics, including those commonly used in population genetics [83], experimental evolution [11, 12, 26], and gene knockout screens [17–19, 42, 43, 80, 81, 106].

However, as we compare these statistics in simulated competition experiments, we find that the choice of time scale can lead to different mutant rankings (Figs. 2B and 4A,B), confound apparent epistasis between mutations (Fig. 3B), and alter the apparent long-term trend in fitness (Fig. 3F), while the choice of the reference subpopulation can lead to a systematic shift in inferred fitness values (Fig. 5C). The quantitative differences between relative fitness statistics affect our ability to make evolutionary predictions. Fixation probabilities and other population genetics quantities depend on ratios of parameters defining the rates of selection, mutation, and genetic drift processes [66, 67]. These rates must be measured in the same units. For example, in batch cultures with serial transfer, genetic drift is set by the dilution bottle-neck and thus operates per cycle (e.g., 1/N per cycle, where N is the bottleneck population size). Thus quantifying a mutation’s relative fitness s (equivalent to the rate of selection) per generation and comparing to genetic drift per cycle (Ns) would incorrectly predict the fixation probability, especially in cases where per-generation and per-cycle fitness rank mutants differently. The choice of statistic may also affect the outcome of a significance test in the context of gene essentiality tests (e.g., using log or logit with the test statistic in Wetmore et al. [18]). Since these measurements often serve as a first screen to narrow down the investigation to the top set of genes [17, 40], a difference in mutant ranking means the investigation might miss relevant genes because of the choice of fitness statistic.

Best practices for quantifying mutant fitness in high-throughput experiments

Based on these insights, we recommend the following choices for quantifying relative fitness of a mutant:

Use the logit encoding of relative abundance, because under the null model of logistic dynamics this linearizes the trajectory of relative abundance. Only use a different encoding if alternate, non-logistic dynamics are known.

For batch growth or any experiment with non-steady state growth, use a fixed extrinsic time scale (e.g., a single growth cycle), rather than an intrinsic time scale (e.g., the number of generations), which introduces additional variation between competitions. Only quantify fitness per generation for experiments that are known to have steady-state growth (e.g., in a chemostat).

In bulk competition experiments, always use the pairwise fitness relative to the wild-type (by including a barcoded wild-type or by grouping neutral mutants into a virtual wild-type) because this more closely matches the relative fitness in pairwise competition.

To control the error from higher-order interactions, we recommend minimizing relative abundance of the mutant library as practically feasible. For a given error tolerance, Eq. (11) gives an estimate for the maximum library abundance.

Our recommendations agree with previous criticisms of relative fitness per-generation [5, 8, 9, 94] and total relative fitness [59, 107] but differ from the standard practice in high-throughput genetic screens [18, 19, 80] and many evolution experiments [33, 91], including the LTEE [11, 26]. The choice of the reference group is a well-recognized issue in high-throughput evolutionary studies, in which one first estimates a total relative fitness and then subtracts a correction based on the fitness of the wild-type [12, 50, 52] or a mean population fitness [13, 41, 42, 51, 107]. In contrast, we recommend to choose the reference group at the level of relative abundance (see Methods for a step-by-step demonstration). While researchers must ultimately choose how to quantify fitness based on the specific goals and conditions of their experiment, we believe that a more consistent adherence to the recommendations above would be valuable to the field as a whole so that fitness data from one study can be quantitatively compared to fitness from another.

How general is the disagreement between fitness statistics in pairwise competition?

The discrepancy in mutant ranking for relative fitness per-cycle scyclelogit and per-generation sgenlogit requires positive covariation between the wild-type and mutant log fold-changes (LFCs) across a set of pairwise competition experiments. This condition is not dependent on any particular mechanism or model, although our minimal model of population dynamics (Fig. 2A) demonstrates one way this covariation can occur: when mutants have diminished yields and are present at sufficiently-high abundances to affect wild-type LFCs (Fig. S7).

Our model focused on the effect of resource competition on fitness discrepancies since that interaction is almost always present between microbial genotypes, providing a baseline expectation. Moreover, resource competition is likely to be the only interaction in high-throughput fitness measurements of mutants with only small genetic differences (e.g., point mutations or gene knockouts) relative to a wild-type. We expect that in cases with additional interactions between mutants and the wild-type, there will be further variation in wild-type LFCs across competitions and thus even more discrepancies betwen fitness statistics. For example, mutants that stimulate wild-type growth via the secretion of nutrients should create positive LFC covariation across mutants and thus be ranked differently per-cycle and per-generation. Indeed, our reanalysis of LTEE competition data shows that fitness per-cycle scyclelogit and pergeneration sgenlogit (equivalent to the statistic W=1+sgenlogit used in the LTEE) disagree on the fitness ranking at almost every time point (Fig. 4A) and in the first 50,000 generations overall (Fig. 4B).

How general is the deviation of bulk fitness statistics from pairwise competition experiments?

The offset in total relative fitness (between grey and orange points in Fig. 5C), equivalent to the mean library fitness (Eq. (S88) in Sec. S13), is independent from the population dynamics (Fig. 5B, Sec. S13). In contrast, the error between pairwise fitness in bulk competition and in pairwise competition (orange points in Fig. 5C) is specific to the population dynamics studied here, although it points to more general mechanisms. Specifically, we expect pairwise fitness in bulk competition to deviate from pairwise competition as soon as the mutant library is able to influence the chemical environment of the competition, for example, when a large fraction of mutants fails to complete respiration of the main carbon source and instead turns to fermentation [108–110], with the release of pathway intermediates like acetate that indirectly lower the pH in the growth medium [111, 112].

The choice of fitness statistic in the light of other sources of discrepancy

Besides the conceptual choices of fitness quantification discussed in this article, experimental limitations can create discrepancies between replicate fitness measurements of the same mutant [36, 50, 113]. For example, relative abundance measurements entail sampling uncertainties (when we sample liquid for colony counting or DNA sequencing) [44, 59, 60], copy number variation [18, 44], as well as PCR jackpots and sequencing read errors [13, 104, 105]. Furthermore, there will inevitably be some variation in initial condition between replicates and fluctuations during the fitness assay [36, 105, 113].

Methods

Inferring growth traits for the single-gene knockout collection in yeast

We use a previously published data set [29] for the single-gene knockout collection in Saccharomyces cerevisiae [99], where the authors track growth of each genotype for 47 hours in monoculture, using microwell plates with defined growth medium. We download this growth curve data set from the PROPHECY database [114], choosing specifically the data set measured in Synthetic Defined medium. We last accessed the PROPHECY website (http://prophecy.lundberg.gu.se/) on March 30, 2020, but as of November 15, 2025 the website is no longer accessible. We have included a reformatted version of the data with our code repository (https://github.com/justuswfink/24FitnessQuantification). From the raw time series of optical density, the authors subtracted a background correction based on blank wells, applied another correction for nonlinearity at high optical densities, and smoothed the growth curve to remove electrical noise [29].

We start with the curves in the published data set (9951 curves) and apply further trimming and smoothing steps (Sec. S6). From this data we calculate the time series of instantaneous per-capita growth rate N−1dN/dt (where N is the optical density) and identify time windows where the rate is approximately constant, which we interpret as distinct growth phases (Sec. S6). We only include curves that have a single phase of constant exponential growth followed by a stationary phase of approximately zero growth (9424 curves).

We quantify the biomass yield, maximum growth rate, and the lag time from each remaining curve as follows. First, we estimate the initial abundance Ninitial (average optical density over first three time points) and the final abundance Nfinal (average optical density over stationary phase) from the growth curve. Then we calculate the biomass yield as Y = (Nfinal − Ninitial)/R(0), where R(0) = 111 mM (20 g/L) is the initial concentration of glucose (assuming glucose is the single limiting resource). To estimate the maximum growth rate, we average the instantaneous growth rate over the exponential phase. Finally, we estimate the lag time from the intersection of the log initial abundance (log N0) with the slope of the maximum growth rate during the exponential growth phase [84]. We excluded curves with negative initial OD, curves with negative inferred lag times, and curves with a low quality of fit between the measured time series and a simulated curve based on the inferred trait values (R2 < 0.95; see below for the model). Our final database includes trait estimates for 9195 curves, which represents 92.4% of the original data set [29].

Each single-gene deletion strain was measured in two technical replicate growth curves, using a second plate with identical layout in the same plate reader. Some genotypes have only one estimate of the growth traits because the other replicate did not pass our filters (273 genotypes), but for most genotypes we retain two replicate estimates in our final data set (4163 genotypes); a few genotypes even have three (2 genotypes) or four replicate estimates (54 genotypes) because these geno-types were included multiple times by the original authors [29]. From these replicates we finally calculate the average yield, growth rate, and lag time for each geno-type. The data set also contained many replicate growth curves of the wild-type strain, 374 of which passed our filters. Since wild-type traits inferred from these replicates had large variation (potentially due to measuring the wild-type across many different plates and days), we define the wild-type trait as the median, not the mean, of these replicates. Although the wild-type measurements included in this data set have large trait variation (orange dots in Fig. S4B,C), the fact that the variation across knockouts is significantly greater than for the wild-type replicates in terms of growth rate (Levene’s test p = 1.3 × 10−9) and lag time (Levene’s test p = 0.025), and that the replicate measurements of gene knockouts are correlated (Fig. S5), suggests that these trait measurements do capture true genetic variation. In contrast, the variation in biomass yield across knockouts is not significantly different from the variation across wild-type replicates (Levene’s test p = 0.44), suggesting that the variation in biomass yield is driven by non-genetic factors like the abundance or batch of resources used for each growth measurement.

Simulating population dynamics in competition experiments

Throughout this work, we model the competition between a set of genotypes during a single batch culture growth cycle using the following model of population dynamics [97, 98]: (12a) 1NjdNjdt=gj(t,R),Nj(0)=xj(0)N0 (12b) dRdt=−∑j1YjdNjdt,R(0)=R0 where for each genotype j, Nj is the absolute abundance, xj(0) the initial relative abundance, gj(t, R) the growth rate, and Yj the biomass yield, where the genotypes include a wild-type as well as one or more mutants. The initial absolute abundance of all genotypes together is N0, and the concentration of the single limiting resource is R with initial value R0. Note that Eq. (12b) assumes that cells consume resources only for biomass growth and not for maintenance of existing biomass.

To capture the sigmoidal shape of typical growth curves with a lag phase, exponential growth phase, and saturation phase (e.g., Fig. S4A), we model the dependence of growth rate on time and resource concentration as (13) gj(t,R)=gjΘ(t−λj)Θ(R), where Θ is the step function: (14) Θ(z)={0z≤01z>0. In this model, a genotype has an initial lag phase of time εj where no growth occurs, followed by a phase of constant exponential growth at rate gj, and ending when the resource concentration R reaches zero. The time to re-source depletion, which we also call the saturation time, is defined by the implicit equation R(tsat) = 0. The simple form of the growth response (Eq. (13)) means that the absolute abundance of genotype j at saturation is (15) Nj(tsat)=Nj(t0)egj(tsat−λj)Θ(tsat−λj) and its log fold-change is (16) LFCj=gj(tsat−λj)Θ(tsat−λj). To calculate relative fitness per-cycle (Eq. (9) and per-generation (Eq. (10)), we numerically determine the saturation time tsat by integrating the biomass and resource dynamics (Eq. (12)) as described in previous work [97, 98]. We note that it also possible to derive an approximate expression for the saturation time tsat which we use for our analytical calculations (Sec. S7).

Testing epistasis between mutations in population dynamics model

Consider a wild-type strain with growth rate gwt = 1, lag time εwt = 2, and biomass yield Ywt = 1. We define six basic mutation that represent the six directions of the trait space in our model: a mutation that increases growth rate (gmut = 1.2; symbol +g), a mutation that decreases growth rate (gmut = 0.8; symbol − g), a mutation that increases lag time (εmut = 3; symbol +ε), a mutation that decreases lag time (εmut = 1; symbol − ε), a mutation that increases yield (Ymut = 4; symbol +Y), and a mutation that decreases yield (Ymut = 0.25; symbol −Y). We combine these mutations into double mutants as follows: mutations are additive at the level of traits (e.g., the double mutant (+g, −g) is neutral) and mutations in different traits are independent. For a given pair of mutations, epistasis is defined as [100] (17) E=s(A,B)−s(A)−s(B), where s(A) is the relative fitness of the single mutant with mutation A, s(B) is the relative fitness of the single mutant with mutation B, and s(A, B) is the relative fitness of the double mutant with both A and B. We estimate the fitness of each mutant by simulating one growth cycle of competition with the wild-type (at initially equal abundance) using an initial biomass of N0 = 0.01 and initial resource concentration R0 = 0.5. These conditions are chosen such that mutations that increase biomass yield reach a maximum density of up to N (tsat) = 2, on the same scale as the optical density measured for the empirical knockouts of S. cerevisiaea (Fig. S4C).

Quantifying total and pairwise relative fitness in bulk competition experiments

We define total and pairwise relative fitness,two alternative measures of fitness in bulk competition experiments, as follows. Consider a population with multiple genotypes such that each genotype i has absolute abundance Ni(t) at time t. The total relative abundance of the genotype is (18) xi(t)=Ni(t)∑jNj(t). The “total” here refers to the fact that this is the abundance of genotype i relative to all other genotypes in the population. Similarly, we can think of the definition of relative fitness discussed in the main text (Eq. (4)) as the total relative fitness of this genotype under an encoding m: (19) σim=ddtm(xi(t)). However, sometimes we want to track the dynamics of a genotype relative to another specific genotype; for example, to follow the fate of a mutant against the wild-type in a bulk competition experiment. For a pair of genotypes i ≠ j, we thus define the pairwise relative abundance (20) xij(t)=xi(t)xi(t)+xj(t), which is the relative abundance of genotype i in a sub-population of only genotypes i and j. To predict the change in the pairwise relative abundance, we can use the slope (21) sijm=ddtm(xij(t)), which defines the pairwise relative fitness of genotype I with respect to genotype j (see Sec. for a generalize definition that covers the case i = j). In the special case of a population with only two strains, the pair-wise relative fitness and total relative fitness are identical (σ2m=s21m) but they may differ with more than two genotypes (Sec. S13).

It is also possible to define the total and pairwise relative fitness statistics as finite differences over a time interval, rather than as instantaneous derivatives. For example, we can define them over a growth cycle starting at t = 0 and ending at t = tsat: (22) σi,cyclem=m(xi(tsat))−m(xi(0)), (23) sij,cyclem=m(xij(tsat))−m(xij(0)). For these fitness statistics in the bulk competitions, we use the logit encoding (24) σi,cyclelogit=log⁡xi(tsat)1−xi(tsat)−log⁡xi(0)1−xi(0), (25) sij,cyclelogit=log⁡xi(tsat)xi(0)−log⁡xj(tsat)xj(0), where we have rewritten sij,cycle in the form that it is presented in bulk competition experiments [15, 51, 52]. The logit encoding has mathematical advantages for coarse-graining the relative fitness of genotype groups (Sec. S11) but using the log encoding is another common choice in the literature [17, 41, 50]. However, the relative abundances of the individual mutants in our simulated bulk competition experiments are low enough that these two encodings are approximately equivalent (logit x ≈ log x for x ≪ 1).

Unlike the comparison between per-cycle and pergeneration relative fitness where we focused on rank differences (Fig. 2B), here we can evaluate absolute differences in fitness estimates because the total and pairwise fitness in bulk fitness are measured in the same units as the relative fitness in pairwise competition. For all mu-tant genotypes, we calculate the absolute error between these bulk fitness estimates (Eqs. (24) and (25)) and the relative fitness in pairwise competitions (Eq. (9)), which we take as the ground truth (Fig. 5C). Note that the pairwise competition could depend on the initial relative abundance of the mutant; we have chosen a very low relative abundance (10−6) that mimics a mutant arising de novo and where this dependence is very weak.

Clearly, minimizing the abundance of the mutant library minimizes the strength of higher-order interactions (also reducing ranking disagreement, Fig. S21B,C). As low mutant abundances create practical problems for experiments, it is valuable to identify a maximum abundance for the mutant library that keeps the error from bulk fitness estimates below a desired threshold (Eq. (11)). It is convenient to express this threshold in relative rather than absolute terms (compare Fig. S23 to Fig. 5C).

A guide to calculate pairwise relative fitness under the logit encoding from bulk competition data

Here ri(t) represents the best estimate of read numbers for lineage i after correcting for experimental artifacts, like jackpot events in the sequencing (e.g., see discussion in Levy et al. [13]).

Calculate the total relative abundance xi of each lineage i by dividing with the total number of reads: (26) xi(t)=ri(t)∑i=1nri(t), where the sum is over all n lineages in the experiment.

Define a group of lineages i = 1, …, k as the reference (e.g., a group of neutral mutants) and sum their total relative abundance (27) xref(t)=∑i=1kxi(t).

Calculate the pairwise relative abundance xi,ref of each lineage by (28) xi,ref(t)=xi(t)xi(t)+xref(t).

Calculate the pairwise fitness per-cycle for each strain i under the logit encoding (29) si,reflogit=1Δt[log⁡xi,ref(t2)1−xi,ref(t2)−log⁡xi,ref(t1)1−xi,ref(t1)], where Δt is the number of growth cycles between time points t1 and t2.